---
title: "Bilingualism in Canada: Geographic Distribution"
author: Esther Schott (esther.schott@mail.concordia.ca)
output: html_document
---
output: 
  html_document: 
    code_folding: show
    collapsed: no
    df_print: kable
    highlight: espresso
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\\\
# load libraries
```{r}
#install.packages("gpclib", type="source")

library(tidyverse)
library(rgeos)
library(rgdal)
library(maptools)
library(sp)
library(here)
library(fuzzyjoin)
library(viridis)
library(scales)
library(ggrepel)
library(rio)
```


# load map data
## census area map data
```{r}


# decent, uncluttered map theme (needs devtools package tho)
devtools::source_gist("https://gist.github.com/hrbrmstr/33baa3a79c5cfef0f6df")

if(!file.exists(here("02_map_data","census_map_2016","lcma000a16a_e.shp"))) {
# grab the file from "Statistics Canada"
# 2016 census, area areas only
download.file("http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/files-fichiers/2016/lcma000a16a_e.zip", destfile=here("02_map_data/census_map_2016.zip"))
unzip(here("02_map_data/census_map_2016.zip"))
system("ogr2ogr canada.shp lcma000a16a_e.shp -simplify 0.01")
}
canada2016 <- readOGR(here("02_map_data","census_map_2016","lcma000a16a_e.shp"),"lcma000a16a_e")





# do this to see what's available from an "identifier" standpoint
# "CDNAME" seems to be the census district name
# "PRNAME" seems to be the province name
# str(canada@data)




map_areas2016 <- data.frame(id=canada2016@data$CMANAME,
                        area=sapply(slot(canada2016, "polygons"), slot, "area") )
cma_map2016 <- fortify(canada2016, region="CMANAME")
rm(canada2016)
canada_map_areas_2016 <- merge(cma_map2016, map_areas2016, by="id")




```
## read census province map
```{r}
if(!file.exists(here("02_map_data","province_map_2016","lpr_000b16a_e.shp"))) {
download.file("http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/files-fichiers/2016/lpr_000b16a_e.zip", destfile=here("02_map_data/lpr_000a16a_e.zip"))
#download.file("http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/files-fichiers/2016/lpr_000a16a_e.zip",
 #             destfile=here("02_map_data/lpr_000a16a_e.zip"))

unzip(here("02_map_data/lpr_000a16a_e.zip"), exdir = here("02_map_data","province_map_2016"))


# this simplifies the polygons so they load/plot faster
system("ogr2ogr  lpr_000b16a_e.shp -simplify 0.01")
}

# what layers do we have? you can use this to check
#   ogrListLayers(here("02_map_data","province_map_2016","lpr_000b16a_e.shp"))
# but there are none, so the shapefile is the layer

province2016 <- readOGR(here("02_map_data","province_map_2016","lpr_000b16a_e.shp"),"lpr_000b16a_e")
province_map <- fortify(province2016, region="PRENAME")
rm(province2016)



```


# process census language data
## load census language data
```{r}
idata_child = import(file = here::here("03_output","00_processed_data", "idata_preprocessed.Rdata")) %>%
  filter(age_group %in% "0 to 4 years")

```

## summarize census language data by area & province
```{r}
(gma_LangGroup_data = idata_child %>%
  group_by(province,area,n_languages) %>%
  summarize(N=n()) %>%
    mutate( Percent = round(N/sum(N),2)) %>% 
    filter(n_languages == "Two or more") )
    

```
# table for poster
```{r}
table_prov = idata_child %>%
  group_by(province) %>%
  mutate(total = n()) %>% 
  group_by(province,total,n_languages) %>%
  summarize(N=n()) %>% 
    mutate( Percent = percent(N/total,1))%>% 
    filter(n_languages != "One Language")  %>%
  select(-n_languages)

table_city = idata_child %>%
  group_by(province,area,n_languages) %>%
  summarize(N=n()) %>%
  
    mutate( Percent_city = percent(N/sum(N),1)) %>%
  ungroup() %>%
mutate(      area = ifelse(area =="Other census metropolitan areas, census agglomerations and o","_other",as.character(area)))%>% 
    filter(n_languages != "One Language")  %>%
  select(-n_languages) %>%
  arrange(area,-N) %>% filter(N >250 ) 


```



## process "other" non-cma area data
```{r}

province_data = gma_LangGroup_data %>%    filter(substr(area,1,5) == "Other")

province_data$province= as.character(province_data$province)

northern_data = data.frame(province= rep("Northern Canada",3), id = c("Yukon", "Northwest Territories", "Nunavut"))

province_data  = province_data %>% full_join(northern_data) %>%
  mutate(id = ifelse(is.na(id), as.character(province), as.character(id)))

```
## save data for poster table
```{r}
table_otherCMA = province_data %>% group_by(province) %>% slice(1) %>% ungroup() %>% select(-id, -N) %>%
  mutate(Percent = percent(Percent, 1))


table_poster = table_prov %>% full_join(table_city %>% 
  rename( Percent = Percent_city))  %>% 
  full_join(table_otherCMA) %>%
  select(province, area,total, N, Percent) %>%
  arrange(province)
write.csv(table_poster, file=here("03_output","Table.csv"),row.names = F)
```

## match area data to map data
```{r}
# remove "other" areas (saved in province_data)
gma_LangGroup_data = gma_LangGroup_data %>% 
   filter(substr(area,1,5) != "Other")

# try matching the census area names to the census map data
attempted_matching = stringdist_left_join(gma_LangGroup_data, 
                             map_areas2016 %>% 
                               rename(area = id), 
                             by ="area", 
                             distance_col = NULL)

# write failures to file to match by hand
attempted_matching %>% select(province,area.x, area.y) %>% 
  # keep only those where matching failed
  filter(is.na(area.y)) %>% 
  # write to file to match by hand
  write.csv(., file=here("02_map_data/Dict-area.csv"), row.names=F) 

# read in the matched-by-hand-codes
dict.map  = read.csv(here("02_map_data/Dict-area-coded.csv"))

# add the hand-coded labels to the dataframe
gma_LangGroup_data = gma_LangGroup_data %>% 
  left_join(dict.map) %>% 
  rename(area.old = area) %>% 
  mutate(area = ifelse(is.na(area.splitUp), area.old, as.character(area.splitUp))) %>%
  select(-area.old, -area.splitUp)

# 
gma_lang_for_merging = gma_LangGroup_data %>% 
    stringdist_left_join(map_areas2016 %>% rename(area = id), 
                         by = "area", distance_col = NULL) %>% 
    select(-area.old, -area.x, -area) %>%
    rename(id = area.y)
```
# merge census data and map data
## CMAs
```{r}
gma_map_data = left_join(gma_lang_for_merging, canada_map_areas_2016)

```

## province data
```{r}
province_map_data = full_join( 
                      province_data %>% 
                        subset(area =="Other census metropolitan areas, census agglomerations and o") , province_map, 
                      by=c("province" ="id"))

```

## get labels

```{r}
gma_labels = gma_map_data %>% group_by(area.old) %>% summarize(
  # y
  long = mean(long), 
  # x 
  lat = mean(lat)) %>%
  mutate(short_label = gsub(" – .*","",area.old),
         short_label = gsub(" / .*","", short_label))

add_labels = tibble(area.old = "Grand Sudbury", lat = 1250604.3, long = 7085883 )
GTA_cities = c("Brantford – Guelph – Barrie","Kitchener – Cambridge – Waterloo",  "Hamilton",  "Kingston – Peterborough", 
 "St. Catharines – Niagara", "Toronto")
gma_labels = gma_map_data %>% group_by(area.old) %>% summarize(
  # y
  long = mean(long), 
  # x 
  lat = mean(lat)) %>%
  full_join(add_labels) %>%
  mutate(short_label = gsub(" – .*","",area.old),
         short_label = gsub(" / .*","", short_label))


gma_labels_adj = gma_labels %>%
  mutate(city = ifelse(area.old == "Ottawa – Gatineau", "Ottawa", area.old ),
         lat = case_when(city == "Ottawa"~ lat+ 100000, 
                         city == "Oshawa" ~lat+ 50000,
                         TRUE ~ lat),
         
    long_adj = case_when(
    city %in% c("Toronto", "Edmonton","Calgary", "Ottawa")~ long + 100000,
    TRUE ~ long),
    lat_adj  = ifelse(city == "Vancouver", lat - 150000, lat),
    GTA = case_when(
      area.old %in% GTA_cities[1:2] ~ "GTA-south",
      area.old %in% GTA_cities ~ "GTA-west",
      TRUE ~NA_character_),
    label_linebreak = gsub(" – ","\n",area.old)
    )
  

# choose which labels to plot
good_labels = 
c("Calgary", "Edmonton",  
"Halifax",  "Kelowna – Abbotsford", "Kingston – Peterborough", 
"Kitchener – Cambridge – Waterloo", "London", "Moncton – Saint John", 
"Montréal", "Ottawa – Gatineau", "Québec", "Regina – Saskatoon", 
  "Toronto", 
"Vancouver", "Victoria", "Windsor", "Winnipeg")

```


# plot area and provinces together
```{r}
province_map_data = province_map_data %>% na.omit()
map_plot =  ggplot() + 
  # draw province map
  geom_map(data=province_data , 
           map=province_map,
           aes(map_id=id, fill = Percent),
           color="white", 
        
           size=0.1)+ 
  # draw area areas
geom_map(data=gma_map_data, 
        map=canada_map_areas_2016,
       aes(map_id=id,    
          group=id, 
         fill=Percent),
         color="white", 
        size=0.05)+ 
  # English label
   scale_fill_viridis(option = "D", labels = scales::percent_format(accuracy=1), 
                      name="Rate of \nchild \nbilingualism", breaks = seq(0.0,0.25, by = .05))+
  #French label (comment out if not wanted)
 # scale_fill_viridis(option = "D", labels = scales::percent_format(accuracy=1), 
  #                    name="Pourcentage des \nenfants \nbilingues", breaks = seq(0.0,0.25, by = .05))+
  # define limits for the plot (since province_map and canada_map_areas uses same scale, can use either one...)
   expand_limits(x = province_map$long, 
                 y = province_map$lat) + theme_map(base_size = 15)+ coord_fixed(ratio=1)+ 
   theme(legend.position = c(0.71, 0.53))
 
 map_plot_with_label = map_plot +   geom_label_repel( aes(long, lat,label = city, group=NULL),
              size=4,
           data = gma_labels_adj) 
 ggsave(here::here("03_output","01_figures","canada_map.png"), width = 10, height =7)
```

# try a petter way to plot labels
```{r}

map_plot + 
  # Montreal, Ottawa,Toronto
geom_text_repel(data          = subset(gma_labels_adj, long > 5000000),
                    aes(x= long, y= lat, label = city, group=NULL),
   nudge_x       = 9195936-  subset(gma_labels_adj, long > 5000000)$long,
    segment.size  = 1,
    segment.color = "grey50",
    direction     = "both",
    vjust         = 1
  ) +
  geom_text_repel(data          = subset(gma_labels_adj, long < 5000000),
                    aes(x= long, y= lat, label = city, group=NULL),
    nudge_y       = 950425 - subset(gma_labels_adj, long < 5000000)$lat,
    segment.size  = 1,
    segment.color = "grey50",
    direction     = "both",
    vjust         = 1
  ) 

 ggsave(here::here("03_output","01_figures","canada_map_diff_labels.png"), width = 13, height =7)



```
##zoom in on ontario + Quebec
```{r fig.height = 7, fig.width = 10}

zoom_plot_x_min = 6971050-10000
zoom_plot_x_max = 7757551+500000
zoom_plot_y_min = 720614.5-10000
zoom_plot_y_max = 1439061.2+200000


gma_labels_adj = gma_labels_adj %>%
  mutate(zoom_area = if_else(between(long,
                                                zoom_plot_x_min,
                                                zoom_plot_x_max)&
                                            between(lat, zoom_plot_y_min,
                                                    zoom_plot_y_max),
                             "zoom", NA_character_))


ratio = (zoom_plot_x_max- zoom_plot_x_min ) /(zoom_plot_y_max-zoom_plot_y_min)
height =7


map_plot + coord_cartesian(xlim=c(zoom_plot_x_min,zoom_plot_x_max),
                           ylim = c(zoom_plot_y_min,zoom_plot_y_max))+
  
  geom_label_repel( aes(long, lat,label = label_linebreak, group=NULL),
              
           data = gma_labels_adj %>% filter(zoom_area == "zoom" & 
                                              is.na(GTA)), 
           alpha=.75,
           size=5, 
           segment.colour = "grey50", segment.size = 1.5)+
  
  geom_label_repel( aes(long, lat,label = label_linebreak, group=NULL),
              
           data = gma_labels_adj %>% filter(zoom_area == "zoom" & 
                                              GTA =="GTA-south"), 
           alpha=.75,ylim= c(817632.7,NA), xlim=c(NA,7026050),
           size=4, 
           segment.colour = "grey50", segment.size = 1.5 #, nudge_x = 50000,nudge_y =-20000
           )   +
  geom_label_repel( aes(long, lat,label = label_linebreak, group=NULL),
              
           data = gma_labels_adj %>% filter(zoom_area == "zoom" & 
                                              GTA =="GTA-west"), 
           alpha=.75,xlim= c(7300000-20000,NA),
           size=5, 
           segment.colour = "grey50", segment.size = 1.5 #, nudge_x = 50000,nudge_y =-20000
           )   +
  guides(fill = "none")



 ggsave(here::here("03_output","01_figures","canada_map_zoomed.png"), width = ratio*height, height =height)

```


# child bilingualism map
```{r}
cutoff= 7041271

essential_gma_labels = gma_labels %>% filter(short_label %in% c("Calgary", "Edmonton", "Halifax", "Regina", "Vancouver", 
"Victoria", "Winnipeg",
                                                              "Montréal", # comment out for poster map
                                                              "Toronto")) # comment out for poster map
map_plot + 
  # Montreal, Ottawa,Toronto
geom_label_repel(data          = subset(essential_gma_labels,  long > cutoff),
                    aes(x= long, y= lat, label = short_label, group=NULL),
   nudge_x       = 9195936-  subset(essential_gma_labels, long > cutoff)$long,
    segment.size  = 1,
    segment.color = "grey50",
    direction     = "both",
    vjust         = 1
  ) +
  geom_label_repel(data          = subset(essential_gma_labels, long < cutoff),
                    aes(x= long, y= lat, label = short_label, group=NULL),
    nudge_y       = 950425 - subset(essential_gma_labels, long < cutoff)$lat,
    segment.size  = 1,
    segment.color = "grey50",
    direction     = "both",
    vjust         = 1
  ) 

 ggsave(here::here("03_output","01_figures","canada_map_most_labels_en.png"), width = 13, height =7)

```


# best canada map
```{r}
good_labels = c( "Calgary", "Edmonton",  "Halifax", "Kelowna",   
 "Regina", 
  "Vancouver", "Victoria", 
"Winnipeg")

map_plot + 
  # Montreal, Ottawa,Toronto
geom_label_repel(data          = subset(gma_labels, short_label %in% good_labels ),
                    aes(x= long, y= lat, label = short_label), size =2 ,
                 alpha = 0.75 , seed=1234)
  
 ggsave(here::here("03_output","01_figures","canada_map_transparent_labels.png"), width = 9, height =7)

```


