---
title: "Census Bilingualism Pre-processing HOUSEHOLDS"
output:
  html_document:
    df_print: paged
---
author: Esther Schott
please email esther.schott@mail.concordia.ca if you have any questions
see poster created from a previous version of this script: https://osf.io/4gekn/

# 1. Preparation Steps
## Script settings
```{r}
load_data_from_raw = TRUE # If true, loads data from .sav, else loads R file (faster)
# needs to be loaded from raw the first time you execute this script on a new computer! 

use_reduced_data_set = FALSE # If true, uses a smaller subset of full data set (for troubleshooting etc.)
```

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(tidylog)
library(rio) # for reading in SPSS files
library(labelled)
library(janitor)
library(codebook)
#library(scales)

```



#  HOUSEHOLDS dataset

## read in dataset 
if this is the first time you use this script on a computer, download the dataset from the odesi database. https://lib-ezproxy.concordia.ca/login?url=https://odesi.ca/ 
Search 2016 Census on odesi, click on the save icon and then download the .sav FILE. Might work better on chrome than firefox. (look out for the "if the download does not start, click here" link that might appear.)
if `load_data_from_raw == FALSE`, we skip this step to save time
```{r}
# need to change to reading in factor solution as above
if(load_data_from_raw == TRUE) {
  raw_hhdata = rio::import(here("01_datasets/pumf-98M0002-E-2016-hierarchical_F1.sav"),format = "sav", user_na = c("Not available", "88888888", "99999999"))
  saveRDS(raw_hhdata, here("03_output/00_processed_data/hdata_raw.Rdata"))} else {
  raw_hhdata = readRDS(here("03_output/00_processed_data/hdata_raw.Rdata"))
}

#dict = tibble(old_name = names(raw_hhdata), explanation = var_label(raw_hhdata))
#export(dict, file = here("01_datasets/ColNamesDict.csv"))
hh_column_names = import(here("01_datasets/ColNamesDict_hhdata.csv"))




```
## data cleaning - general

### clean up column names
```{r}
# we only want to rename columns that we created a new name for
hh_column_names_keep = hh_column_names %>%
  filter(new_name!="")
# rename columns to new name
# pull() is necessary to convert column to vector (which rename_at expects)
raw_hhdata = raw_hhdata %>% rename_at(vars(pull(hh_column_names_keep,old_name)), ~ pull(hh_column_names_keep,new_name))
```
### convert to categorical variables to factors to see variable labels
```{r}
# make factor - the spss variable has all variables stored as numeric, which makes them read into R as numeric (which means it doesn't show labels present in the dataset). Can use factorize to remedy that 

# from the column dictionary, keep only columns that are categorical 
columns_to_convert_to_factor = hh_column_names %>%
  filter(type == "categorical") %>%
  mutate(current_name = if_else(new_name== "", old_name, new_name)) %>% # some of these columns have been renamed, use new name if it was renamed
  pull(current_name)


# batch convert to factor
# this takes a bit long
raw_hhdata = raw_hhdata %>%
  mutate_at(.vars = vars(columns_to_convert_to_factor),
             .funs = factorize)
# now, instead of numbers e.g. for provice we see the actual label in the df
```



### select variables of interest
```{r}
# select which variables are necessary for our analysis
# keep all that we have renamed (to include more, edit the ColNamesDict.csv in datasets folder)

hhdata = raw_hhdata %>% select(pull(hh_column_names_keep, new_name))
```

### create variable for households with children 9 and under

```{r}
# create a variable that says whether child younger than 10 lives in household


hhdata = hhdata %>% 
  mutate(child0_9 = ifelse(age_group == "0 to 9 years", 1,0)) %>%
    group_by(household_id) %>%
  mutate(children_per_hh = sum(child0_9)) %>%
  ungroup()



```

### use subset of full data (for troubleshooting)
This is necessary while creating the code because some code is SLOW. change use_reduced_data_set to change at the top of the script
```{r}

if(use_reduced_data_set == TRUE) {
  # select X unique households to keep in subset
  household_subsample = sample(unique(hhdata$household_id),
                               20000)  
  hhdata = hhdata %>%
    filter(household_id %in% household_subsample)
}

```

### clean up area variable
Change "Other census metropolitan areas, Census Agglomerations and [..]" to just "other"
```{r}
hhdata = hhdata %>%
  mutate(area = sub(" .*", "", area)) # delete all strings after a space

```



## data cleaning -   language variables
### shorten variable levels
```{r}

# shorten the variable levels to just have true/false
# from "True - Respondent reported English [..]" to just "True"
hhdata = hhdata %>%
  mutate_at( .vars = vars(lang_most_eng, lang_most_fre, lang_regular_eng, lang_regular_fre, lang_regular_oth),
             ~ fct_relabel(., ~str_remove(., '\\s*-.*'))) # remove all characters after space


```
# create true/false variable for lang_most_other
This is the question where people write in their language, and we also want a true/false vector for this question so it paralells all the other language variables
```{r}
# create a variable that functions like all other home language variables and just says true/false for lang_most_oth_write_in  
hhdata = hhdata %>% 
  mutate(lang_most_oth = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "False",
    lang_most_oth_write_in == "Not available" ~ NA_character_,
    !is.na(lang_most_oth_write_in) ~ "True"
  ))
# test that this is working: 
# table(hhdata$lang_most_oth_write_in, hhdata$lang_most_oth)


```



### clean up Write-in component for non-official languages
clean up so variable empty if no other language spoken, NA if no info available, and has the name of the language if there is another language spoken
```{r}
hhdata = hhdata %>%
  
  mutate(lang_most_oth_write_in = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "",
    lang_most_oth_write_in == "Not available" ~ NA_character_,
    lang_most_oth_write_in == "All other single languages" ~ "Other",
    # in all other cases, shorten the answer to just the first word, for easier reference
    TRUE ~  sub(" .*", "", lang_most_oth_write_in)
  ))
    
```



### calculate number of languages spoken (individual)

```{r}
# for each of these columns, we want to know if the answer is True 
true_false_language_columns = c("lang_most_eng", "lang_most_fre", "lang_most_oth","lang_regular_eng", "lang_regular_fre", "lang_regular_oth")


# count how many languages each person speaks:
# if the answer to any of the above columns is True, count them
hhdata = hhdata  %>% 
  mutate(N_languages_person_numeric = rowSums(.[true_false_language_columns] == "True", na.rm = TRUE),
         N_languages_person = # make this a categorical variable
           case_when(N_languages_person_numeric == 0 ~ NA_character_,
                     N_languages_person_numeric == 1 ~ "One Language",
                     N_languages_person_numeric > 1 ~ "Two or more",
                     TRUE ~ NA_character_))

# check result:
#View(hhdata %>% select(household_id, person_id, lang_most_eng:lang_regular_oth, lang_most_oth, N_languages_individual))


```

### create language pairs variable

```{r}
# we need a column that says what languages that person speaks. So English_French, English&Italian&Other etc. 
# if a person speaks a language at home (value is "True"), then copy the column name to new column lang_pair_ordered

# sorted by most spoken then regularly spoken languages (for the person)
hhdata$lang_pair_ordered <- apply(hhdata[true_false_language_columns], 1, function(x) paste(names(x[x=="True"]), collapse="&"))



# this takes a long time to run 
# now we need to clean up the messy column names in each row:
hhdata = hhdata %>%
  mutate( lang_pair_ordered = str_remove(lang_pair_ordered, "(NA&)+|(&NA)+"), # remove NA in beginning and middle
   lang_pair_ordered =str_remove(lang_pair_ordered,"&NA"), # remove NAs at End
   lang_pair_ordered = str_replace(lang_pair_ordered, "(lang_most_eng)+|(lang_regular_eng)+", "English"), # substite English column names with word english
   lang_pair_ordered = str_replace(lang_pair_ordered, "(lang_most_fre)+|(lang_regular_fre)+", "French"), # same for french
   lang_pair_ordered = str_replace(lang_pair_ordered, "lang_most_oth", as.character(lang_most_oth_write_in)),
   lang_pair_ordered = str_replace(lang_pair_ordered, "lang_regular_oth", "Other") # for lang_regular_oth we don't know language, just call it other
   )

# check result:
#View(hhdata %>% select(household_id, person_id, lang_most_eng:lang_regular_oth, lang_most_oth, N_languages_individual, lang_pair_ordered))




```
### create variables for each language in dataset
```{r}

hhdata  =  hhdata %>%
  separate(lang_pair_ordered, c("lang1", "lang2", "lang3", "lang4"), remove = FALSE)
# warning that not all people have 4 languages is normal
```



### simplify language pairs variable (sorted A-Z)
```{r}

# remove order by most spoken then regularly spoken languages (now sorted by A-Z)
hhdata$lang_pair <- apply(hhdata[c("lang1", "lang2", "lang3","lang4")], 1, function(x) paste(sort(x[!is.na(x)]), collapse="&"))

```
## get columns back in alphabetical order
```{r}
hhdata = hhdata %>% select(sort(names(hhdata)))
```


# Descriptives
## keep only households with kids under 10
```{r}
# restrict to kids under 10
# keep only households that have at least one child under 10
hhdata_child = hhdata %>%
  filter(children_per_hh >0)
```


## summary tables: children under 9 years
```{r}


# number of bilinguals
#canadian average:
hhdata_child %>%
filter(age_group=="0 to 9 years") %>%
  group_by(N_languages_person) %>%
  summarize(N=n()) %>%
    mutate( Percent = N/sum(N)) 





(area_lang_data = hhdata_child %>%
filter(age_group=="0 to 9 years") %>%
  group_by(province,area,N_languages_person) %>%
  summarize(N=n()) %>%
    mutate( Percent = round(N/sum(N),3))%>%
    filter(N_languages_person != "One Language"))


# # specific language combos
 hhdata_child_langPair = hhdata_child %>%
   filter(age_group=="0 to 9 years",N_languages_person_numeric>1 ) %>%
   group_by(province,lang_pair) %>%
   summarize(N=n())%>%
   mutate( Percent = round(N/sum(N),3))%>%
   arrange(province, -N)

 hhdata_child_langPair=hhdata_child_langPair %>% group_by(province) %>%
   slice(1:5) %>%
   mutate(id=seq_along(Percent))

# add nicer language labels to graph (need to adapt to new data)
#lang_dict = read_csv(file= here("01_datasets","LangPair_Dict_2016.csv"))
#hhdata_child_langPair = full_join(hhdata_child_langPair, lang_dict, by = c("lang_pair"= "Lang_Combo_top2" )


```


# export for analysis in other scripts
```{r}

# this contains ALL households (in case another analysis needs all data)
export(hhdata, file = here("03_output/00_processed_data/hhdata_preprocessed.Rdata"))
```





# getting data from only adults in child's households
```{r}
# filter out children
hhdata_child_adult <- hhdata_child %>% 
  filter(child0_9 == 0) %>% 
  filter(age_group != "10 to 14 years" & age_group != "15 to 19 years")
  
# see number of adults per household
adults_n <- hhdata_child_adult %>% 
  # group by household
  group_by(household_id) %>% 
  summarize(n_adults = n())
# number of adults per household ranges 1 - 6



### language

# get language (pairs) and number of speakers per household
adults_lang <- hhdata_child_adult %>% 
  group_by(household_id, lang_pair_ordered) %>% 
  summarize(n_lang_pair = n())

# get number of bilingual adults per household
adults_biling_n <- adults_lang %>% 
  filter(str_detect(lang_pair_ordered, "&")) %>% 
  group_by(household_id) %>% 
  summarize(n_biling_adults = sum(n_lang_pair)) 

# get number of monolingual adults per household
adults_monoling_n <- adults_lang %>% 
  filter(!str_detect(lang_pair_ordered, "&")) %>% 
  mutate(lang_pair_ordered = na_if(lang_pair_ordered, "NA")) %>% 
  filter(!is.na(lang_pair_ordered)) %>% 
  group_by(household_id) %>% 
  summarize(n_monoling_adults = sum(n_lang_pair))

# combine info on adults' language
adults_n <- adults_n %>% 
  left_join(adults_biling_n, by = "household_id") %>% 
  left_join(adults_monoling_n, by = "household_id") %>% 
  # add values in if we know there are no reported bilingual adults in the HH
  mutate(n_biling_adults = case_when(
    n_adults - n_monoling_adults == 0 ~ 0,
    TRUE ~ as.numeric(as.character(.$n_biling_adults))
  )) %>% 
  mutate(n_biling_adults_group = case_when(
           n_biling_adults == 0 ~ "None",
           n_biling_adults == 1 ~ "One Adult",
           n_biling_adults == 2 ~ "Two Adults",
           n_biling_adults > 2 ~ "Three or More Adults"
          )) %>% 
  mutate(n_biling_adults_group = as.factor(n_biling_adults_group)) 

adults_n$n_biling_adults_group <- factor(adults_n$n_biling_adults_group, levels = c("None", "One Adult", "Two Adults", "Three or More Adults"))

# # get proportion of bilingual adults per household
# adults_n <- adults_n %>% 
#   left_join(adults_biling_n, by = "household_id") %>% 
#   mutate(n_biling_adults = replace_na(n_biling_adults, 0), 
#          prop_biling_adults = n_biling_adults / n_adults,
#          
         # create factor for groups of # of bilingual adults


### education

# currently 8 factor levels. want to change to 5 to match Turcotte analyses. merging post-secondary levels together

hhdata_child_adult <- hhdata_child_adult %>% 
  mutate(education_rank = case_when(
    education == "No certificate, diploma or degree" ~ "No high school diploma",
    education == "Secondary (high) school diploma or equivalency certificate" ~ "High school diploma",
    education ==  "Trades certificate or diploma other than Certificate of Appr" |
    education ==  "Certificate of Apprenticeship or Certificate of Qualificatio" |
    education ==  "College, CEGEP or other non-university certificate or diplom" |
    education ==  "University certificate or diploma below bachelor level"  ~ "Post-secondary diploma",
    education == "Bachelor's degree" ~ "Bachelor's degree", 
    education == "University certificate, diploma or degree above bachelor lev" ~ "Graduate degree"
  ))

  # make an ordered factor
hhdata_child_adult$education_rank  <-  factor(hhdata_child_adult$education_rank, levels = c("No high school diploma", "High school diploma", "Post-secondary diploma", "Bachelor's degree", "Graduate degree"), ordered = TRUE)

adults_edu <- hhdata_child_adult %>% 
  group_by(household_id) %>% 
  summarize(highest_ed = max(education_rank)) %>% 
  mutate(highest_ed = as.factor(highest_ed))



```


# create df for only children

```{r}
hhdata_child_only <- hhdata_child %>% 
  # filter for only children
  filter(child0_9 == 1) %>% 
  # join with # of bilingual adults
  left_join(adults_n, by = "household_id") %>% 
  # join with highest education
  left_join(adults_edu, by = "household_id") %>% 
  ### income - creating quintiles, following Turcotte analyses
  mutate(income_quintile = case_when(
    income_decile == "In first decile" | income_decile == "In second decile" ~ "In first quintile",
    income_decile == "In third decile" | income_decile == "In fourth decile" ~ "In second quintile",
    income_decile == "In fifth decile" | income_decile == "In sixth decile" ~ "In third quintile",
    income_decile == "In seventh decile" | income_decile == "In eighth decile" ~ "In fourth quintile",
    income_decile == "In ninth decile" | income_decile == "In tenth decile" ~ "In fifth quintile",
  )) %>% 
  mutate(income_quintile = as.factor(income_quintile))

hhdata_child_only$income_quintile <- factor(hhdata_child_only$income_quintile, levels = c("In first quintile", "In second quintile", "In third quintile", "In fourth quintile", "In fifth quintile"))
```


# Export child only data for predictor analysis in separate script
``` {r}
export(hhdata_child_only, file = here::here("03_output","00_processed_data", "hhdata_child_only.Rdata")) 
```