---
title: "Census Bilingualism Analysis"
output:
  html_document:
    df_print: paged
---
author: Esther Schott
please email esther.schott@mail.concordia.ca if you have any questions
see poster created from this script: https://osf.io/4gekn/

# 1. Preparation Steps
## Script settings
```{r}
load_data_from_raw = FALSE # If true, loads data from .sav, else loads R file (faster)

use_reduced_data_set = TRUE # If true, uses a smaller subset of full data set (for troubleshooting etc.)
```

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(rio) 
library(labelled)
library(janitor)
library(codebook)

```




# INDIVIDUALS dataset

###  read in invididuals file

```{r}




# 2016 data files
# need to update to most recent
if(load_data_from_raw == TRUE) {
raw_idata = import(here("01_datasets/pumf-98M0001-E-2016-individuals_F1.sav"), 
                      format = "sav")
saveRDS(raw_idata, here("03_output/00_processed_data/idata_raw.Rdata"))} else {
raw_idata =   readRDS(here("03_output/00_processed_data/idata_raw.Rdata"))
}


# convert strings to factors 
idata <- lapply(raw_idata, 
                    Int2Factor)
# convert list to dataframe
idata <- as.data.frame(idata, 
                           stringsAsFactors = FALSE)

# select which variables are necessary for our analysis
variables.keep = c("PPSORT", # individual case ID
  "PR","CMA", # location: province & metropolitain area
                   "AGEGRP","PKID0_1", "PKID2_5", # number of children between 0 and 1, and 2 and 5
                   "HLAEN","HLAFR","HLANO", # home language part A
                   "HLBEN","HLBFR","HLBNO", #home language part B
                    "CFInc", "DPGRSUM")
idata_short = idata %>% select(variables.keep)
idata_short$Child0_1 = factor(idata_short$PKID0_1, labels=c("NotAvailable","1+","0"))
idata_short$Child2_5 = factor(idata_short$PKID2_5, labels=c("NotAvailable","1+","0"))

# create age_bracket variable... 
idata_short$age_bracket = ifelse(as.integer(idata_short$AGEGRP)<16, "adult","child")

idata_short = idata_short %>%
  rename(caseID = "PPSORT")


```
## use subset of full data (for troubleshooting)
```{r}
if(use_reduced_data_set == TRUE) {
  idata_short = idata_short %>%
    sample_n(50000)
}
```


###  change factor levels & variable names for language variables
this needs to be cleaned up & shortened (Esther)
```{r}



# change factor levels for question part a ("language used most often")
idata_short$HLA_En = factor(idata_short$HLAEN, labels= c(NA,T,F))
levels(idata_short$HLA_En) <- list(NotAvailable = NA,English = "TRUE",N="FALSE")
levels(idata_short$HLA_En)[levels(idata_short$HLA_En)=='NotAvailable'] <- NA


idata_short$HLA_Fr = factor(idata_short$HLAFR, labels= c(NA,T,F))
levels(idata_short$HLA_Fr) <- list(NotAvailable =NA, French = "TRUE",N="FALSE")
levels(idata_short$HLA_Fr)[levels(idata_short$HLA_Fr)=='NotAvailable'] <- NA


idata_short$HLA_Oth = factor(idata_short$HLANO)
levels(idata_short$HLA_Oth)[levels(idata_short$HLA_Oth)=='No non-official language'] <- "N"
levels(idata_short$HLA_Oth)[levels(idata_short$HLA_Oth)== 'Not available'] <- NA

levels(idata_short$HLA_Oth)[levels(idata_short$HLA_Oth)=='All other languages'] <- "Other"

# change for question part b (regular use)
idata_short$HLB_En = factor(idata_short$HLBEN, labels= c(NA, T,F))
levels(idata_short$HLB_En) <- list(NotAvailable =NA, English = "TRUE",N="FALSE")
levels(idata_short$HLB_En)[levels(idata_short$HLB_En)=='NotAvailable'] <- NA

idata_short$HLB_Fr = factor(idata_short$HLBFR, labels= c(NA,T,F))
levels(idata_short$HLB_Fr) <- list(NotAvailable =NA, French = "TRUE",N="FALSE")
levels(idata_short$HLB_Fr)[levels(idata_short$HLB_Fr)=='NotAvailable'] <- NA


idata_short$HLB_Oth = factor(idata_short$HLBNO, labels = c(T, F))
levels(idata_short$HLB_Oth) <- list(Other = "TRUE",N="FALSE")
#levels(idata_short$HLB_Oth)[levels(idata_short$HLB_Oth)=='N'] <- NA


```

### convert from wide to long
For easier counting of languages etc.
```{r}
old_language_cols = c("HLAEN","HLAFR", "HLANO","HLBEN",  "HLBFR", "HLBNO" )
idata_long = idata_short %>% 
  select(-old_language_cols) %>%
  # cut down size of df for debugging
  #sample_n(193000) %>%

  gather(lang_question, answer,HLA_En:HLB_Oth) %>%
  arrange(caseID)


```

#### create bilingualism variable (wide df)
```{r}
idata_long = idata_long %>%
 mutate(language_reported = if_else(answer =="N",0,1)) %>%
  group_by(caseID) %>%
  mutate(N_Lang = sum(language_reported, na.rm=T),
         N_Two_or_more = case_when(
                                  N_Lang == 1 ~ "1",
                                  N_Lang >1 ~ "2+",
                                  is.na(N_Lang) ~ NA_character_),
        # Lang_Combo_ranked = paste(answer[language_reported==1], 
        #                           collapse ="_"),
        # Lang_Combo_a_z = paste(sort(answer[language_reported==1]), 
        #                        collapse ="_"),
         Lang_Combo_top2 = paste(sort(answer[language_reported==1][1:2]), 
                                 collapse ="_")) 

idata_long = idata_long %>% ungroup()
# idata_long %>% filter(N_Lang>1) %>% View()

```

#### get back to wide (one-case-one-row)
```{r}
idata_wide = idata_long %>% select(-lang_question, -answer, -language_reported) %>%
  distinct()
```



### create variable for geographical locations of interest
```{r}
idata_long$gma.interest = ifelse(idata_long$CMA %in% c("Vancouver", "Montréal", "Toronto","Ottawa - Gatineau" , "Calgary", "Edmonton"),as.character(idata_long$CMA), NA)
```


 

## keep only children 0-4, save idata_child for creating map


```{r}

idata_child = idata_wide %>%
  filter(AGEGRP == "0 to 4 years")
saveRDS(idata_child, file = here::here("03_output","00_processed_data", "Individuals_child_data.rdata"))

  
#gma_LangGroup_data = gma_LangGroup_data %>% 
#  mutate(Area = agrep()

```




#  Individuals dataset: descriptives
### summary tables: children across canada
check how distributions over province and metropolitain areas are
```{r}
# select only adults who are living with at least one child
idata_hh_child = idata_wide %>% 
  filter(PKID0_1 =="One or more" | PKID2_5 =="One or more") %>% 
  filter(age_bracket=="adult")

# how many are there per province & area?
idata_hh_child %>% group_by(PR, CMA) %>%
  summarize(N=n()) %>% 
  spread(PR, N)

# breakdown by province
idata_child %>% group_by(PR) %>%
  summarize(N=n())

# some provinces have  few data points. can probably concentrate on BC, Alberta, Ontario, Quebec



# top metropolitain areas
idata_child %>% group_by(CMA) %>%
   summarize(N=n()) %>%
  arrange(-N)
# top metropolitain areas are: Toronto, Montréal, Vancouver Calgary, Ottawa - Gatineau, Edmonton (cutoff could be 5000 individuals?)

```
## bilingualism in children in individuals dataset
#### summary table: bilingual children across canada

```{r}
# commented out because I need to fix langGroup
# idata_short %>% filter(age_bracket == "child") %>%
#   group_by(AGEGRP, N_languages_individual_cat) %>%
#   summarize(N=n()) %>%
#   spread(LangGroup_individual, N) %>%
#   clean_names() %>%
#   mutate(total = one_language+two_or_more, percent = two_or_more/total)


```

#### summary tables: bilingual children under 6yo by province & metropolitan area
```{r}
# commented out because I need to fix langGroup

# idata_short %>% filter(AGEGRP == "0 to 4 years"| AGEGRP =="5 to 6 years") %>%
#   group_by(gma.interest, LangGroup_individual) %>%
#   summarize(N=n()) %>%
#   spread(LangGroup_individual, N) %>%
#   clean_names() %>%
#   mutate(total = one_language+two_or_more, percent = round(two_or_more/total,3)) 

```


#### summary tables: number of bilingual adults living with a child under 5

```{r}
# commented out because I need to fix langGroup

# 
# idata_child %>%
#   filter(Child0_1 == "1+"  ) %>%
#   group_by(LangGroup_individual) %>%
#   summarize(Count = n())%>%
#   mutate(           Percent = Count/sum(Count))
# 
# # broken down by key metropolotain areas
# 
# idata_child %>%
#   filter(!is.na(gma.interest)& Child0_1 == "1+" |Child2_5=="1+") %>%
#   group_by( gma.interest, LangGroup_individual) %>%
#   summarize(Count = n())%>%
#   mutate(           Percent = Count/sum(Count)) 
#   
# 
# 
# # need to investigate what to do with these NA in other language
# idata_child %>% filter(is.na(LangGroup)) %>% 
#   group_by(LangPair_simplified) %>% 
#   summarize(count=n())
# 
# # look at 2-5 year olds
# idata_child %>% 
#   filter(Child2_5 == "1+") %>%
#   group_by(LangGroupBil) %>%
#   summarize(Count = n())  %>%
#   mutate(           Percent = Count/sum(Count))

# these look relatively the same as the one-year olds, focus on the one-year olds becuase they are our population of interest and 2-5 year olds likely get more exposure outside of the house (not captured in this survey)
```
