---
title: "Bilingualism in Canada: Geographic Distribution"
author: Esther Schott (esther.schott@mail.concordia.ca)
output: html_document
---
output: 
  html_document: 
    code_folding: show
    collapsed: no
    df_print: kable
    highlight: espresso
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Readme
important term: CMA = census metropolitan area. can be a city (e.g., Montreal) or a combination of two cities that are in the same province (if both cities are small, can be geographically close or not). In this script, I use city/metropolitan area/cma interchangeably

the map for the census paper is composed of three main components:
- map data for provinces (matched with the non-metropolitain areas % for each province). This provides the background for the plot
- map data for CMAs (matched with % for CMAs). overlaid on the non-cma % per province
- labels for CMAs (not all are shown in the current iteration of the map to prevent clutter)

# load libraries
```{r load packages}

library(tidyverse)
library(rgeos) #map-related stuff
library(rgdal) #map-related stuff
library(maptools) #map-related stuff
library(sp) #map-related stuff
library(here) # create working directory
library(rio) # easier loading & saving
library(geojsonio)
library(spdplyr) # the `dplyr` counterpart for shapes
library(rmapshaper) # the package that allows geo shape transformation
```







## load census metropolitain area map data
CMA = census metropolitain area. can be a city (e.g., Montreal) or a combination of two cities that are in the same province (if both cities are small, can be geographically close or not)

the map files are downloaded from the census website. If you have any questions to what the files contain, there is an explanation pdf that comes with the download. 
```{r load_cma_maps}



if(!file.exists(here("02_map_data","census_map_2016","lcma000a16a_e.shp"))) {
# grab the file from "Statistics Canada"
# 2016 census, area areas only
# see the pdf in map_data/census_map_2016 for more info
download.file("http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/files-fichiers/2016/lcma000b16a_e.zip", destfile=here("02_map_data/census_map_2016.zip"))
unzip(zipfile = here("02_map_data/census_map_2016.zip"), exdir =  here("02_map_data/census_map_2016"))

}

# what layers do we have? you can use this to check
#   ogrListLayers(here("02_map_data","province_map_2016","lpr_000b16a_e.shp"))
# but there are none, so the shapefile is the layer

canada_census_2016_cma <- readOGR(dsn = here("02_map_data","census_map_2016",
                            "lcma000b16a_e.shp"),
                      layer = "lcma000b16a_e",
                      use_iconv = TRUE, 
                      encoding = "UTF-8")



canada_census_2016_cma = ms_simplify(canada_census_2016_cma, sys = TRUE, keep= .1)


# this throws error message, this seems useful to fix this in future: https://stackoverflow.com/questions/46003060/error-using-region-argument-when-fortifying-shapefile

cma_coordinates <- fortify(canada_census_2016_cma, region="CMANAME")%>% 
  rename(area_maps = "id")

rm(canada_census_2016_cma)
cma_coordinates <- cma_coordinates %>%
  #merge(cma_coordinates, map_areas2016, by="area_maps") %>%
  rename(id = area_maps) %>%
  relocate(long, lat, .before = id)



```
### simplify
```{r}
canada_raw_json <- geojson_json(canada_raw)
# simplify the coastlines so the map is easier to plot (see here for more info: https://cran.r-project.org/web/packages/rmapshaper/vignettes/rmapshaper.html) 
# this takes forever!
canada_census_2016_cma = ms_simplify(canada_census_2016_cma, sys = TRUE)

```


## read census province map

Useful  info on canadian map data in R here: https://tengl.net/blog/2020/1/7/drawing-canada-maps-in-r
```{r read_province_maps}
if(!file.exists(here("02_map_data","province_map_2016","lpr_000b16a_e.shp"))) {
download.file("http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/files-fichiers/2016/lpr_000b16a_e.zip", destfile=here("02_map_data/lpr_000a16a_e.zip"))
#download.file("http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/files-fichiers/2016/lpr_000a16a_e.zip",
 #             destfile=here("02_map_data/lpr_000a16a_e.zip"))

unzip(here("02_map_data/lpr_000a16a_e.zip"), exdir = here("02_map_data","province_map_2016"))


# this simplifies the polygons so they load/plot faster
}

# what layers do we have? you can use this to check
#   ogrListLayers(here("02_map_data","province_map_2016","lpr_000b16a_e.shp"))
# but there are none, so the shapefile is the layer

canada_raw <- readOGR(here::here("02_map_data",
                             "province_map_2016",
                             "lpr_000b16a_e.shp"),
                        layer = "lpr_000b16a_e",
                        use_iconv = TRUE, encoding = "UTF-8")




# remove big map object to ease load on memory 
rm(province_census_2016)




```


##Simplify data
```{r}
canada_raw_json <- geojson_json(canada_raw)
# simplify the coastlines so the map is easier to plot (see here for more info: https://cran.r-project.org/web/packages/rmapshaper/vignettes/rmapshaper.html) 
# this takes forever!
canada_raw_sim <- ms_simplify(canada_raw_json, sys = TRUE) # 3

```

## save output
```{r}
geojson_write(canada_raw_sim, file = "data/canada_cd_sim.geojson") # 4

```


