---
title: "Census Bilingualism Analysis"
output:
  html_document:
    df_print: paged
---
author: Esther Schott
please email esther.schott@mail.concordia.ca if you have any questions
see poster created from a previous version of this script: https://osf.io/4gekn/

# 1. Preparation Steps
## Script settings
```{r}
load_data_from_raw = FALSE # If true, loads data from .sav, else loads R file (faster)

use_reduced_data_set = FALSE # If true, uses a smaller subset of full data set (for troubleshooting etc.)
```

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(rio) # for reading in SPSS files
library(labelled)
#library(cowplot)
#library(viridis)
library(janitor)

library(codebook)


library(scales)

```



#  HOUSEHOLDS dataset

## To Dos:
- some variables have SPSS style missing values (9999999), but I don't think we're using any of them?
- need to clean up lang_most_oth_write_in (remove  language names in parentheses, make labels less wordy...)
- need to create simpified lang_pair_ordered variable that does not take order into account (treats English_French and French_English the same) - but maybe this is only needed for plotting and not for regression?
```{r}

```


## read in dataset 
```{r}
# need to change to reading in factor solution as above
if(load_data_from_raw == TRUE) {
  raw_hhdata = rio::import(here("01_datasets/pumf-98M0002-E-2016-hierarchical_F1.sav"),format = "sav")
  saveRDS(raw_hhdata, here("03_output/00_processed_data/hdata_raw.Rdata"))} else {
  raw_hhdata = readRDS(here("03_output/00_processed_data/hdata_raw.Rdata"))
}

#dict = tibble(old_name = names(raw_hhdata), explanation = var_label(raw_hhdata))
#export(dict, file = here("01_datasets/ColNamesDict.csv"))
hh_column_names = import(here("01_datasets/ColNamesDict.csv"))




```
## data cleaning - general

### clean up column names
```{r}
# we only want to rename columns that we created a new name for
hh_column_names_keep = hh_column_names %>%
  filter(new_name!="")
# rename columns to new name
# pull() is necessary to convert column to vector (which rename_at expects)
raw_hhdata = raw_hhdata %>% rename_at(vars(pull(hh_column_names_keep,old_name)), ~ pull(hh_column_names_keep,new_name))
```
### convert to categorical variables to factors to see variable labels
```{r}
# make factor - the spss variable has all variables stored as numeric, which makes them read into R as numeric (which means it doesn't show labels present in the dataset). Can use factorize to remedy that 

# from the column dictionary, keep only columns that are categorical 
columns_to_convert_to_factor = hh_column_names %>%
  filter(type == "categorical") %>%
  mutate(current_name = if_else(new_name== "", old_name, new_name)) %>% # some of these columns have been renamed, use new name if it was renamed
  pull(current_name)


# batch convert to factor
# this takes a bit long
raw_hhdata = raw_hhdata %>%
  mutate_at(.vars = vars(columns_to_convert_to_factor),
             .funs = factorize)
# now, instead of numbers e.g. for provice we see the actual label in the df
```



### select variables of interest
```{r}
# select which variables are necessary for our analysis
# keep all that we have renamed (to include more, edit the ColNamesDict.csv in datasets folder)

hhdata = raw_hhdata %>% select(pull(hh_column_names_keep, new_name))
```





### create variable for households with children 9 and under

```{r}
# create a variable that says whether child younger than 10 lives in household


hhdata = hhdata %>% 
  mutate(child0_9 = ifelse(age_group == "0 to 9 years", 1,0)) %>%
    group_by(household_id) %>%
  mutate(children_per_hh = sum(child0_9)) %>%
  ungroup()



```

### use subset of full data (for troubleshooting)
This is necessary while creating the code because some code is SLOW. change use_reduced_data_set to change at the top of the script
```{r}

if(use_reduced_data_set == TRUE) {
  # select X unique households to keep in subset
  household_subsample = sample(unique(hhdata$household_id),
                               20000)  
  hhdata = hhdata %>%
    filter(household_id %in% household_subsample)
}

```

### clean up area variable
```{r}
hhdata = hhdata %>%
  mutate(area = sub(" .*", "", area))

```



## data cleaning -   language variables
```{r}

#make this simpler


hhdata = hhdata %>%
  mutate_at( .vars = vars(lang_most_eng, lang_most_fre, lang_regular_eng, lang_regular_fre, lang_regular_oth),
             ~ fct_relabel(., ~str_remove(., '\\s*-.*')))

# create a variable that functions like all other home language variables and just says true/false for lang_most_oth_write_in  
hhdata = hhdata %>% 
  mutate(lang_most_oth = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "False",
    lang_most_oth_write_in == "Not available" ~ NA_character_,
    !is.na(lang_most_oth_write_in) ~ "True"
  ))
# test that this is working: 
# table(hhdata$lang_most_oth_write_in, hhdata$lang_most_oth)



```
### clean up Write-in component for non-official languages
```{r}
hhdata = hhdata %>%
  
  mutate(lang_most_oth_write_in = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "",
    lang_most_oth_write_in == "Not available" ~ NA_character_,
    lang_most_oth_write_in == "All other single languages" ~ "Other",
    TRUE ~  sub(" .*", "", lang_most_oth_write_in)
  ))
    
```



### calculate number of languages spoken (individual)

```{r}
# count how many lanuages each person speaks:
true_false_language_columns = c("lang_most_eng", "lang_most_fre", "lang_most_oth","lang_regular_eng", "lang_regular_fre", "lang_regular_oth")

hhdata = hhdata  %>% 
  mutate(N_languages_person_numeric = rowSums(.[true_false_language_columns] == "True", na.rm = TRUE),
         N_languages_person = # make this a categorical variable
           case_when(N_languages_person_numeric == 0 ~ NA_character_,
                     N_languages_person_numeric == 1 ~ "One Language",
                     N_languages_person_numeric > 1 ~ "Two or more",
                     TRUE ~ NA_character_))

# check result:
#View(hhdata %>% select(household_id, person_id, lang_most_eng:lang_regular_oth, lang_most_oth, N_languages_individual))


```

### create language pairs variable

```{r}
# we need a column that says what languages that person speaks. So English_French, English&Italian&Other etc. 
# if a person speaks a language at home (value is "True"), then copy the column name to new column lang_pair_ordered

# sorted by most spoken then regularly spoken languages (for the person)
hhdata$lang_pair_ordered <- apply(hhdata[true_false_language_columns], 1, function(x) paste(names(x[x=="True"]), collapse="&"))



# this takes a long time to run 
# now we need to clean up the messy column names in each row:
hhdata = hhdata %>%
  mutate( lang_pair_ordered = str_remove(lang_pair_ordered, "(NA&)+|(&NA)+"), # remove NA in beginning and middle
   lang_pair_ordered =str_remove(lang_pair_ordered,"&NA"), # remove NAs at End
   lang_pair_ordered = str_replace(lang_pair_ordered, "(lang_most_eng)+|(lang_regular_eng)+", "English"), # substite English column names with word english
   lang_pair_ordered = str_replace(lang_pair_ordered, "(lang_most_fre)+|(lang_regular_fre)+", "French"), # same for french
   lang_pair_ordered = str_replace(lang_pair_ordered, "lang_most_oth", as.character(lang_most_oth_write_in)),
   lang_pair_ordered = str_replace(lang_pair_ordered, "lang_regular_oth", "Other") # for lang_regular_oth we don't know language, just call it other
   )

# check result:
#View(hhdata %>% select(household_id, person_id, lang_most_eng:lang_regular_oth, lang_most_oth, N_languages_individual, lang_pair_ordered))




```
### create variables for each language in dataset
```{r}

hhdata  =  hhdata %>%
  separate(lang_pair_ordered, c("lang1", "lang2", "lang3", "lang4"), remove = FALSE)
# warning that not all people have 4 languages is normal
```



### simplify language pairs variable (sorted A-Z)
```{r}

# remove order by most spoken then regularly spoken languages (now sorted by A-Z)
hhdata$lang_pair <- apply(hhdata[c("lang1", "lang2", "lang3","lang4")], 1, function(x) paste(sort(x[!is.na(x)]), collapse="&"))

```
## get columns back in alphabetical order
```{r}
hhdata = hhdata %>% select(sort(names(hhdata)))
```


# Descriptives
## keep only households with kids under 10
```{r}
# restrict to kids under 10
# keep only households that have at least one child under 10
hhdata_child = hhdata %>%
  filter(children_per_hh >0)
```


## summary tables: children under 9 years
```{r}


# number of bilinguals
#canadian average:
hhdata_child %>%
filter(age_group=="0 to 9 years") %>%
  group_by(N_languages_person) %>%
  summarize(N=n()) %>%
    mutate( Percent = N/sum(N)) 





(area_lang_data = hhdata_child %>%
filter(age_group=="0 to 9 years") %>%
  group_by(province,area,N_languages_person) %>%
  summarize(N=n()) %>%
    mutate( Percent = round(N/sum(N),3))%>%
    filter(N_languages_person != "One Language"))


# # specific language combos
 hhdata_child_langPair = hhdata_child %>%
   filter(age_group=="0 to 9 years",N_languages_person_numeric>1 ) %>%
   group_by(province,lang_pair) %>%
   summarize(N=n())%>%
   mutate( Percent = round(N/sum(N),3))%>%
   arrange(province, -N)

 hhdata_child_langPair=hhdata_child_langPair %>% group_by(province) %>%
   slice(1:5) %>%
   mutate(id=seq_along(Percent))

# add nicer language labels to graph (need to adapt to new data)
#lang_dict = read_csv(file= here("01_datasets","LangPair_Dict_2016.csv"))
#hhdata_child_langPair = full_join(hhdata_child_langPair, lang_dict, by = c("lang_pair"= "Lang_Combo_top2" )


```

# export for analysis in other scripts
```{r}

# this contains ALL households (in case another analysis needs all data)
export(hhdata, file = here("03_output/00_processed_data/hhdata_preprocessed.Rdata"))
```



