---
title: "Bilingualism in Canada: Preparation for Map"
author: Esther Schott (esther.schott@mail.concordia.ca)
output: html_document
---
output: 
  html_document: 
    code_folding: show
    collapsed: no
    df_print: kable
    highlight: espresso
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Readme
important term: CMA = census metropolitan area. can be a city (e.g., Montreal) or a combination of two cities that are in the same province (if both cities are small, can be geographically close or not). In this script, I use city/metropolitan area/cma interchangeably

the map for the census paper is composed of three main components:
- map data for provinces (matched with the non-metropolitain areas % for each province). This provides the background for the plot
- map data for CMAs (matched with % for CMAs). overlaid on the non-cma % per province
- labels for CMAs (not all are shown in the current iteration of the map to prevent clutter)

# load libraries
```{r load packages}

library(tidyverse)
library(rgeos) #map-related stuff
library(rgdal) #map-related stuff
library(maptools) #map-related stuff
library(sp) #map-related stuff
library(here) # create working directory
library(rio) # easier loading & saving
library(here)
library(geojsonio)
library(spdplyr) # the `dplyr` counterpart for shapes
library(rmapshaper) # the package that allows geo shape transformation
```

# combined shape file
Useful  info on canadian map data in R here: https://tengl.net/blog/2020/1/7/drawing-canada-maps-in-r

the map files are downloaded from the census website. 
```{r}
unzip(zipfile = here("02_map_data/recombined_shape_files_lambert.zip"), exdir =  here("02_map_data/simplified_census_map_2016_lambert"))

canada_census_2016 <- readOGR(dsn = here("02_map_data","simplified_census_map_2016_lambert",
                            "recombined_shape_files_lambert.shp"),
                      layer = "recombined_shape_files_lambert",
                      use_iconv = TRUE, 
                      encoding = "UTF-8")

st_crs(canada_census_2016) 



# make a simpler dataframe
canada_census_2016_df <- fortify(canada_census_2016, 
                        region="name")
# save for next script
export(canada_census_2016_df, file = here("02_map_data/simplified_map_data.Rdata")) 

```



