---
title: "Census Bilingualism - analyzing predictors of bilingualism"
author: "Lena V. Kremin"
date: "7/7/2021"
output: html_document
---
# Prep script

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(rio)
library(arm)
library(boot)
library(mice)
```


## read in data
```{r}
hhdata_child_only = import(here::here("03_output","00_processed_data", "hhdata_child_only.Rdata"))
```



# exploratory data visualization - # of bilingual children by

## bilingal adults in household (HH)

```{r}
# number of bilingual children by number of bilingual adults in HH (HH without biling adults removed, so more detail can be seen)
hhdata_child_only %>% 
  filter(n_biling_adults != 0) %>% 
ggplot(aes(x = factor(n_biling_adults), fill = N_languages_person)) +
  geom_bar(position = "dodge")

# proportion of bilingual children in HH with various numbers of biling adults
ggplot(hhdata_child_only, aes(x = factor(n_biling_adults), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  geom_abline(intercept = 0.5, slope = 0)

# grouped number of bilingual adults
ggplot(hhdata_child_only, aes(x = factor(n_biling_adults_group), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  geom_abline(intercept = 0.5, slope = 0)

## for only Quebec
hhdata_child_only %>% 
  filter(province == "Quebec") %>% 
  ggplot(aes(x = factor(n_biling_adults_group), fill = N_languages_person)) +
    geom_bar(position = "fill") +
    geom_abline(intercept = 0.5, slope = 0)
```

## province / cities

```{r}
# proportion of bilingual children by province
ggplot(hhdata_child_only, aes(x = province, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))

ggplot(hhdata_child_only, aes(x = province, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90)) + 
  facet_grid(. ~ n_biling_adults)

# proportion of bilingual children in major cities
ggplot(hhdata_child_only, aes(x = area, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))
```

## parental education

```{r}
ggplot(hhdata_child_only, aes(x = factor(highest_ed), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))

# by province
ggplot(hhdata_child_only, aes(x = factor(province), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))+
  facet_grid( . ~ highest_ed)

## get total number of children for each parental education level
n_children_edu <- hhdata_child_only %>% 
  group_by(highest_ed) %>% 
  summarize(n_children_edu = n())
```

## immigration status

```{r}
ggplot(hhdata_child_only, aes(x = immigration_generation, fill = N_languages_person)) +
  geom_bar(position = "fill") 
#+  theme(axis.text.x = element_text(angle = 90))

ggplot(hhdata_child_only, aes(x = immigration_generation, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ province)
```

## gender

```{r}
ggplot(hhdata_child_only, aes(x = sex, fill = N_languages_person)) +
  geom_bar(position = "fill") 
```

## income

```{r} 
ggplot(hhdata_child_only, aes(x = income_decile, fill = N_languages_person)) +
  geom_bar(position = "fill") 

ggplot(hhdata_child_only, aes(x = income_quintile, fill = N_languages_person)) +
  geom_bar(position = "fill")
```


# exploratory descriptive calculations

## number of bilingual kids by bilingual adults in household

```{r}

## get total number of children in the database at number of bilingual adults
n_children_adults <- hhdata_child_only %>% 
  mutate(n_biling_adults = as.factor(n_biling_adults)) %>% 
  group_by(n_biling_adults) %>% 
  summarize(n_children_adults = n())

# percentage of bilingual children based on number of bilingual adults in their HH (O, 1, 2+)

biling_by_adults <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  mutate(n_biling_adults = as.factor(n_biling_adults)) %>% 
  group_by(n_biling_adults) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_adults, by = "n_biling_adults") %>% 
  mutate(percent_children = n_children/n_children_adults * 100)


## get total number of children in the database at each level
n_children_adults_group <- hhdata_child_only %>% 
  group_by(n_biling_adults_group) %>% 
  summarize(n_children_adults_group = n())

# percentage of bilingual children based on number of bilingual adults in their HH (O, 1, 2+)

biling_by_adults_group <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(n_biling_adults_group) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_adults_group, by = "n_biling_adults_group") %>% 
  mutate(percent_children = n_children/n_children_adults_group * 100)

```

## province / cities
```{r}

## get total number of children in each province
n_children_province <- hhdata_child_only %>% 
  group_by(province) %>% 
  summarize(n_children_province = n())

# percentage of bilingual children per province based on number of bilingual adults in their HH (O, 1, 2+)

biling_by_adults_province <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(n_biling_adults_group, province) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_province, by = "province") %>%
  mutate(percent_children = n_children/n_children_province * 100)

```

## parental education

```{r}
# percentage of bilingual children based on highest education of parents
biling_by_edu <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(highest_ed) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_edu, by = "highest_ed") %>%
  mutate(percent_children = n_children/n_children_edu * 100)

## get total number of children for each parental education level in each province
n_children_edu_province <- hhdata_child_only %>% 
  group_by(highest_ed, province) %>% 
  summarize(n_children_edu_province = n())

# percentage of bilingual children based on highest education of parents by province
biling_by_edu_province <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(highest_ed, province) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_edu_province, by = c("highest_ed", "province")) %>%
  mutate(percent_children = n_children/n_children_edu_province * 100)
```

## immigration status

```{r}
## get total number of children in each immigration generation
n_children_immigration <- hhdata_child_only %>% 
  group_by(immigration_generation) %>% 
  summarize(n_children_immigration = n())

# percentage of bilingual children per immigration generation
biling_by_immigration <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(immigration_generation) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_immigration, by = "immigration_generation") %>%
  mutate(percent_children = n_children/n_children_immigration * 100)


## get total number of children for each immigration generation in each province
n_children_immigration_province <- hhdata_child_only %>% 
  group_by(immigration_generation, province) %>% 
  summarize(n_children_immigration_province = n())

# percentage of bilingual children based on immigration generation by province
biling_by_immigration_province <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(immigration_generation, province) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_immigration_province, by = c("immigration_generation", "province")) %>%
  mutate(percent_children = n_children/n_children_immigration_province * 100)
```

## income

```{r}
## get total number of children in each income decile
n_children_income <- hhdata_child_only %>% 
  group_by(income_decile) %>% 
  summarize(n_children_income = n())

# percentage of bilingual children per immigration generation

biling_by_income <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(income_decile) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_income, by = "income_decile") %>%
  mutate(percent_children = n_children/n_children_income * 100)
```


# MODELS** 

## prep data 

```{r}
# select and transform relevant variables
hhdata_model <- hhdata_child_only %>% 
  dplyr::select(person_id, household_id, N_languages_person, province, n_biling_adults_group, highest_ed, income_quintile, immigration_generation, fr_eng) %>% 
  mutate(N_languages_person = as.factor(N_languages_person),
         immigration_generation = na_if(immigration_generation, "Not available"),
         immigration_generation = factor(immigration_generation),
         child_biling = case_when(
           N_languages_person == "One Language" ~ 0,
           N_languages_person == "Two or more" ~ 1
         )) 

# reverse difference code variables #HK note: the info on this page seems to call this kind of coding "Backward Difference Coding" https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#backward
## allows for comparison of mean of given level to mean of previous level; level 1 as intercept

## manually create contrast matrices

### used for number of bilingual adults (for only bilingually exposed children)
contr_diff_3 <- matrix(c(-2/3, 1/3, 1/3, -1/3, -1/3, 2/3), ncol = 2)

### used for number of bilingual adults, immigration generation
contr_diff_4  <-  matrix(c(-3/4, 1/4, 1/4, 1/4, -1/2, -1/2, 1/2, 1/2, -1/4, -1/4, -1/4, 3/4), ncol = 3)

### used for parents' highest education, income quintile
contr_diff_5 <- matrix(c(-4/5, 1/5, 1/5, 1/5, 1/5, -3/5, -3/5, 2/5, 2/5, 2/5, -2/5, -2/5, -2/5, 3/5, 3/5, -1/5, -1/5, -1/5, -1/5, 4/5), ncol = 4)

## assign contrasts
contrasts(hhdata_model$n_biling_adults_group) <- contr_diff_4
contrasts(hhdata_model$highest_ed) <- contr_diff_5
contrasts(hhdata_model$income_quintile) <- contr_diff_5
contrasts(hhdata_model$immigration_generation) <- contr_diff_4


# filter for only quebec data - NOT USED IN ANALYSES REPORTED IN PAPER
hhdata_model_quebec <- hhdata_model %>% 
  filter(province == "Quebec")

# filter for Rest of Canada data - NOT USED IN ANALYSES REPORTED IN PAPER
hhdata_model_roc <- hhdata_model %>% 
  filter(province != "Quebec")


# create exposed df - only those children with at least 1 bilingual adult, because we need to redo contrasts when "none" level is removed

hhdata_model_exp <- hhdata_model %>% 
  filter(n_biling_adults_group != "None") %>% 
  mutate(n_biling_adults_group = factor(n_biling_adults_group))

contrasts(hhdata_model_exp$n_biling_adults_group) <- contr_diff_3

hhdata_model_exp_fr_eng <- hhdata_model_exp %>% 
  filter(fr_eng == TRUE)

hhdata_model_exp_other <- hhdata_model_exp %>% 
  filter(fr_eng == FALSE)

# filter for only quebec data - NOT USED IN ANALYSES REPORTED IN PAPER
hhdata_model_quebec_exp <- hhdata_model_exp %>% 
  filter(province == "Quebec")

# filter for Rest of Canada data - NOT USED IN ANALYSES REPORTED IN PAPER
hhdata_model_roc_exp <- hhdata_model_exp %>% 
  filter(province != "Quebec")
```

## *All of Canada model - Model reported in paper

```{r}
m_all <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model, family = "binomial")

summary(m_all)


# model for bilingually-exposed children (those who have at least 1 bilingual adult in HH)

m_exp <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_exp, family = "binomial")

summary(m_exp)

# model for French-English bilingually-exposed children

m_exp_fr_eng <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_exp_fr_eng, family = "binomial")

summary(m_exp_fr_eng)

# model for non-French-English bilingually-exposed children 

m_exp_other <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_exp_other, family = "binomial")

summary(m_exp_other)
```

## Rest of Canada model - NOT USED IN ANALYSES REPORTED IN PAPER

```{r}
m_roc_all <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_roc, family = "binomial")

summary(m_roc_all)

# model for only bilingually-exposed children 

m_roc_exp <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_roc_exp, family = "binomial")

summary(m_roc_exp)
```

## Quebec model - NOT USED IN ANALYSES REPORTED IN PAPER

```{r}
m_quebec_all <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_quebec, family = "binomial")

summary(m_quebec_all)

# model for only bilingually-exposed children 

m_quebec_exp <- glm(child_biling ~ n_biling_adults_group + highest_ed + income_quintile + immigration_generation, data = hhdata_model_quebec_exp, family = "binomial")

summary(m_quebec_exp)
```

## Model Criticism

```{r}

# Model crticism not reported in paper

# # function
# lrAcc <- function(lrMod, responseVar, use.ranef=TRUE){
#   if(!is.factor(model.frame(lrMod)[,responseVar])){ model.frame(lrMod)[,responseVar] <- as.factor(model.frame(lrMod)[,responseVar]) }
#   if(use.ranef){ preds = predict(lrMod, newdata=model.frame(lrMod)) } else{
#     preds = predict(lrMod, newdata=model.frame(lrMod), re.form=NA) }
#   preds <- ((sign(preds)/2)+0.5)
#   respVarValues <- model.frame(lrMod)[,responseVar]
#   if(is.numeric(respVarValues)){ y <- respVarValues } else{
#     y <- (as.numeric(model.frame(lrMod)[,responseVar])-1) }
#   acc <- sum(preds==y)/length(preds)
#   return(acc)
# }
# 
# 
# # calculate accuracies
# lrAcc(m_roc_all, 'child_biling') #87.51%
# lrAcc(m_roc_exp, as.factor("child_biling")) #58.77%
# lrAcc(m_quebec_all, as.factor("child_biling")) #89.68%
# lrAcc(m_quebec_exp, as.factor("child_biling")) #63.26%

# residual plots

binnedplot(predict(m_roc_all), resid(m_roc_all))
binnedplot(predict(m_roc_exp), resid(m_roc_exp))
binnedplot(predict(m_quebec_all), resid(m_quebec_all))
binnedplot(predict(m_quebec_exp), resid(m_quebec_exp))

# cook's distance

#ROC
roc_cooks <- hhdata_model_roc %>% 
  filter(!is.na(n_biling_adults_group) & !is.na(highest_ed) & !is.na(income_quintile) & !is.na(immigration_generation) & !is.na(child_biling)) %>% 
  mutate(cooksDistance=glm.diag(m_roc_all)$cook)

ggplot(aes(x=cooksDistance), data=roc_cooks) + 
  geom_histogram()

#ROC exp
roc_exp_cooks <- hhdata_model_roc_exp %>% 
  filter(!is.na(n_biling_adults_group) & !is.na(highest_ed) & !is.na(income_quintile) & !is.na(immigration_generation) & !is.na(child_biling)) %>% 
  mutate(cooksDistance=glm.diag(m_roc_exp)$cook)

ggplot(aes(x=cooksDistance), data=roc_exp_cooks) + 
  geom_histogram()

#Quebec
quebec_cooks <- hhdata_model_quebec %>% 
  filter(!is.na(n_biling_adults_group) & !is.na(highest_ed) & !is.na(income_quintile) & !is.na(immigration_generation) & !is.na(child_biling)) %>% 
  mutate(cooksDistance=glm.diag(m_quebec_all)$cook)

ggplot(aes(x=cooksDistance), data=quebec_cooks) + 
  geom_histogram()

#Quebec exp
quebec_exp_cooks <- hhdata_model_quebec_exp %>% 
  filter(!is.na(n_biling_adults_group) & !is.na(highest_ed) & !is.na(income_quintile) & !is.na(immigration_generation) & !is.na(child_biling)) %>% 
  mutate(cooksDistance=glm.diag(m_quebec_exp)$cook)

ggplot(aes(x=cooksDistance), data=quebec_exp_cooks) + 
  geom_histogram()
```

## *Predicted probabilities from model - reported in paper as Table 4

```{r}
# function to convert logit space to probabilities
logit2prob <- function(logit){
  odds <- exp(logit)
  prob <- odds / (1 + odds)
  return(prob)
}

# steps for creating predictions
## 1. create df with all possible combinations of levels of variables
## 2. predict values for each combination
## 3. convert predicted values from logits to probabilities
## 4. average across levels of a single predictor



## 1. create df with all possible combinations of levels of variables

### for all children
newdata_all <- with(hhdata_model, expand.grid(n_biling_adults_group = unique(n_biling_adults_group),
                                              highest_ed = unique(highest_ed),
                                              income_quintile = unique(income_quintile),
                                              immigration_generation = unique(immigration_generation)))

### for only bilingually exposed children
newdata_exp <- with(hhdata_model_exp, expand.grid(n_biling_adults_group = unique(n_biling_adults_group),
                                              highest_ed = unique(highest_ed),
                                              income_quintile = unique(income_quintile),
                                              immigration_generation = unique(immigration_generation)))

### for only Fr-Eng bilingually exposed children
newdata_exp_fr_eng <- with(hhdata_model_exp_fr_eng, expand.grid(n_biling_adults_group = unique(n_biling_adults_group),
                                              highest_ed = unique(highest_ed),
                                              income_quintile = unique(income_quintile),
                                              immigration_generation = unique(immigration_generation)))

### for only other bilingually exposed children
newdata_exp_other <- with(hhdata_model_exp_other, expand.grid(n_biling_adults_group = unique(n_biling_adults_group),
                                              highest_ed = unique(highest_ed),
                                              income_quintile = unique(income_quintile),
                                              immigration_generation = unique(immigration_generation)))

## 2. predict values for each combination

# Canada all
newdata_all$prediction <- predict(m_all, newdata=newdata_all)
canada_all <- newdata_all

# ROC all
newdata_all$prediction <- predict(m_roc_all, newdata=newdata_all)
roc_all <- newdata_all

# Quebec all
newdata_all$prediction <- predict(m_quebec_all, newdata=newdata_all)
quebec_all <- newdata_all


# Canada exposed
newdata_exp$prediction <- predict(m_exp, newdata=newdata_exp)
canada_exp <- newdata_exp

# Canada exposed - Fr-Eng
newdata_exp_fr_eng$prediction <- predict(m_exp_fr_eng, newdata=newdata_exp_fr_eng)
canada_exp_fr_eng <- newdata_exp_fr_eng

# Canada exposed - other
newdata_exp_other$prediction <- predict(m_exp_other, newdata=newdata_exp_other)
canada_exp_other <- newdata_exp_other

# ROC exposed
newdata_exp$prediction <- predict(m_roc_exp, newdata=newdata_exp)
roc_exp <- newdata_exp

# Quebec exposed
newdata_exp$prediction <- predict(m_quebec_exp, newdata=newdata_exp)
quebec_exp <- newdata_exp


## 3. convert predicted values from logits to probabilities using logit2prob function
## 4. average across levels of a single predictor

### function to do steps 3 and 4 
prob_pred <- function(pred){
  pred <- pred %>%
    filter(!is.na(prediction)) %>% 
    mutate(probability = logit2prob(prediction))
  
  adults <- pred %>% 
    group_by(n_biling_adults_group) %>% 
    summarize(mean_prob = mean(probability)) %>% 
    rename(predictor = n_biling_adults_group)
  
  edu <- pred %>% 
    group_by(highest_ed) %>% 
    summarize(mean_prob = mean(probability)) %>% 
    rename(predictor = highest_ed)
  
  income <- pred %>% 
    group_by(income_quintile) %>% 
    summarize(mean_prob = mean(probability)) %>% 
    rename(predictor = income_quintile)
  
  immigration <- pred %>% 
    group_by(immigration_generation) %>% 
    summarize(mean_prob = mean(probability)) %>% 
    rename(predictor = immigration_generation)
  
  rbind(adults, edu, income, immigration)
}

### apply function to each of the 4 models
canada_all_prob <- prob_pred(canada_all)
canada_exp_prob <- prob_pred(canada_exp)
canada_exp_fr_eng_prob <- prob_pred(canada_exp_fr_eng)
canada_exp_other_prob <- prob_pred(canada_exp_other)
roc_all_prob <- prob_pred(roc_all)
roc_exp_prob <- prob_pred(roc_exp)
quebec_all_prob <- prob_pred(quebec_all)
quebec_exp_prob <- prob_pred(quebec_exp)
```

## *Actual percentages from data - reported in paper as Table 3

```{r}
# get the actual percentage of bilingual kids at each predictor level

biling_pct <- function(hhdata_df){
  
  adults <- hhdata_df %>% 
    group_by(child_biling, n_biling_adults_group) %>% 
    summarize(n = n()) %>% 
    pivot_wider(names_from = child_biling, values_from = n) %>% 
    rename(predictor = n_biling_adults_group, monolingual = `0`, bilingual = `1`) %>% 
    filter(!is.na(predictor)) %>% 
    mutate(total_n = monolingual + bilingual,
           biling_pct = bilingual / total_n * 100) %>% 
    dplyr::select(predictor, biling_pct)
  
  edu <- hhdata_df %>% 
    group_by(child_biling, highest_ed) %>% 
    summarize(n = n()) %>% 
    pivot_wider(names_from = child_biling, values_from = n) %>% 
    rename(predictor = highest_ed, monolingual = `0`, bilingual = `1`) %>% 
    filter(!is.na(predictor)) %>% 
    mutate(total_n = monolingual + bilingual,
           biling_pct = bilingual / total_n * 100) %>% 
    dplyr::select(predictor, biling_pct)

  
  income <- hhdata_df %>% 
    group_by(child_biling, income_quintile) %>% 
    summarize(n = n()) %>% 
    pivot_wider(names_from = child_biling, values_from = n) %>% 
    rename(predictor = income_quintile, monolingual = `0`, bilingual = `1`) %>% 
    filter(!is.na(predictor)) %>% 
    mutate(total_n = monolingual + bilingual,
           biling_pct = bilingual / total_n * 100) %>% 
    dplyr::select(predictor, biling_pct)

  
  immigration <- hhdata_df %>% 
    group_by(child_biling, immigration_generation) %>% 
    summarize(n = n()) %>% 
    pivot_wider(names_from = child_biling, values_from = n) %>% 
    rename(predictor = immigration_generation, monolingual = `0`, bilingual = `1`) %>% 
    filter(!is.na(predictor)) %>% 
    mutate(total_n = monolingual + bilingual,
           biling_pct = bilingual / total_n * 100) %>% 
    dplyr::select(predictor, biling_pct)

  
  rbind(adults, edu, income, immigration) 

}

canada_all_pct <- biling_pct(hhdata_model)
canada_exp_pct <- biling_pct(hhdata_model_exp)
canada_exp_fr_eng_pct <- biling_pct(hhdata_model_exp_fr_eng)
canada_exp_other_pct <- biling_pct(hhdata_model_exp_other)
roc_all_pct <- biling_pct(hhdata_model_roc)
roc_exp_pct <- biling_pct(hhdata_model_roc_exp)
quebec_all_pct <- biling_pct(hhdata_model_quebec)
quebec_exp_pct <- biling_pct(hhdata_model_quebec_exp)
```

## reshape data for tables for manuscript

```{r}
# RoC vs Quebec - NOT USED IN FINAL MANUSCRIPT

## actual percentage data
roc_all_table <- roc_all_pct %>% 
  rename(`Rest of Canada` = biling_pct)
roc_exp_table <- roc_exp_pct %>% 
  rename(`Rest of Canada (bilingually-exposed)` = biling_pct)
quebec_all_table <- quebec_all_pct %>% 
  rename(`Quebec` = biling_pct)
quebec_exp_table <- quebec_exp_pct %>% 
  rename(`Quebec (bilingually-exposed)` = biling_pct)

table_pct <- roc_all_table %>% 
  left_join(quebec_all_table, by = "predictor") %>%
  left_join(roc_exp_table, by = "predictor") %>% 
  left_join(quebec_exp_table, by = "predictor") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  add_row(predictor = "Immigration generation", .before = 15) %>% 
  add_row(predictor = "Income quintile", .before = 10) %>% 
  add_row(predictor = "Highest level of parental education", .before = 5) %>% 
  add_row(predictor = "Number of bilingual adults in the household", .before = 1) %>% 
  rename(Characteristic = predictor)

write_csv(table_pct, here("03_output/Table-Percent-Multilingual-By-Characteristic.csv"))


## Canada as a whole - USED IN FINAL MANUSCRIPT

### actual percentage data

canada_all_table <- canada_all_pct %>% 
  rename(`All children` = biling_pct)
canada_exp_table <- canada_exp_pct %>% 
  rename(`Bilingually-exposed children (All)` = biling_pct)
canada_exp_fr_eng_table <- canada_exp_fr_eng_pct %>% 
  rename(`Bilingually-exposed children (English-French)` = biling_pct)
canada_exp_other_table <- canada_exp_other_pct %>% 
  rename(`Bilingually-exposed children (Other language pairs)` = biling_pct)


canada_table_pct <- canada_all_table %>% 
  left_join(canada_exp_table, by = "predictor") %>% 
  left_join(canada_exp_fr_eng_table, by = "predictor") %>% 
  left_join(canada_exp_other_table, by = "predictor") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  add_row(predictor = "Immigration generation", .before = 15) %>% 
  add_row(predictor = "Income quintile", .before = 10) %>% 
  add_row(predictor = "Highest level of parental education", .before = 5) %>% 
  add_row(predictor = "Number of bilingual adults in the household", .before = 1) %>% 
  rename(Characteristic = predictor)

write_csv(canada_table_pct, here("03_output/Table-Percent-Multilingual-By-Characteristic-All-Canada.csv"))


## prediction data
canada_all_table <- canada_all_prob %>% 
  rename(`All children` = mean_prob)
canada_exp_table <- canada_exp_prob %>% 
  rename(`Bilingually-exposed children (All)` = mean_prob)
canada_exp_fr_eng_table <- canada_exp_fr_eng_prob %>% 
  rename(`Bilingually-exposed children (English-French)` = mean_prob)
canada_exp_other_table <- canada_exp_other_prob %>% 
  rename(`Bilingually-exposed children (Other language pairs)` = mean_prob)


canada_table_prob <- canada_all_table %>% 
  left_join(canada_exp_table, by = "predictor") %>% 
  left_join(canada_exp_fr_eng_table, by = "predictor") %>% 
  left_join(canada_exp_other_table, by = "predictor") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  add_row(predictor = "Immigration generation", .before = 15) %>% 
  add_row(predictor = "Income quintile", .before = 10) %>% 
  add_row(predictor = "Highest level of parental education", .before = 5) %>% 
  add_row(predictor = "Number of bilingual adults in the household", .before = 1) %>% 
  rename(Characteristic = predictor)

write_csv(canada_table_prob, here("03_output/Table-Probabilities-All-Canada.csv"))
```

##### missing data investigation

```{r}
missing <-md.pattern(hhdata_model)
missing_roc <- md.pattern(hhdata_model_roc)
missing_roc_exp <- md.pattern(hhdata_model_roc_exp)
missing_quebec <- md.pattern(hhdata_model_quebec)
missing_quebec_exp <- md.pattern(hhdata_model_quebec_exp)
```
We loose data from children missing at least one predictor (19 QA, 23 QE, 34 RA, 42 RE). After those with full data, only missing income is largest group in each model. First gen children and highest ed as high school have higher rates of not reporting income

```{r}
ggplot(hhdata_model, aes(x = immigration_generation, fill = income_quintile)) +
  geom_bar(position = "fill")

ggplot(hhdata_model, aes(x = highest_ed, fill = income_quintile)) +
  geom_bar(position = "fill")

ggplot(hhdata_model, aes(x = n_biling_adults_group, fill = income_quintile)) +
  geom_bar(position = "fill")

hhdata_missing <- hhdata_model %>% 
  mutate(location = case_when(
    province == "Quebec" ~ "Quebec",
    TRUE ~ "RoC"),
    exposed = case_when(
      n_biling_adults_group == "None" ~ "no",
      is.na(n_biling_adults_group) ~ "NA",
      TRUE ~ "yes"
  )) %>% 
  filter(n_biling_adults_group != "NA")

ggplot(hhdata_missing, aes(x = 1, fill = income_quintile)) +
  geom_bar(position = "fill")+
  facet_grid(exposed ~ location)

ggplot(hhdata_missing, aes(x = 1, fill = highest_ed)) +
  geom_bar(position = "fill")+
  facet_grid(exposed ~ location)

ggplot(hhdata_missing, aes(x = 1, fill = immigration_generation)) +
  geom_bar(position = "fill")+
  facet_grid(exposed ~ location)

hhdata_missing %>% 
  filter(location == "RoC" & exposed == "yes") %>% 
  ggplot(aes(x = 1, fill = income_quintile)) +
  geom_bar(position = "fill") +
  facet_grid(immigration_generation ~ n_biling_adults_group)
```

