---
title: "Census Bilingualism Analysis"
output:
  html_document:
    df_print: paged
---
author: Esther Schott
please email esther.schott@mail.concordia.ca if you have any questions
see poster created from this script: https://osf.io/4gekn/

# Preparation Steps
## Script settings
```{r}
load_data_from_raw = TRUE # If true, loads data from .sav, else loads R file (the latter is faster)
# needs to be loaded from raw the first time you execute this script on a new computer!  

use_reduced_data_set = FALSE # If true, uses a smaller subset of full data set (for troubleshooting etc.)
```

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(rio) # for more powerful import & export of data
library(labelled)
library(janitor) # cleaning up variable names
library(codebook) # better documentation
library(snakecase)

```




# INDIVIDUALS dataset

###  read in invididuals file
if this is the first time you use this script on a computer, download the dataset from the odesi database. https://lib-ezproxy.concordia.ca/login?url=https://odesi.ca/ 
Search 2016 Census on odesi, click on the save icon and then download the .sav FILE. Might work better on chrome than firefox. (look out for the "if the download does not start, click here" link that might appear.)
if `load_data_from_raw == FALSE`, we skip this step to save time
```{r}


# 2016 data files
if(load_data_from_raw == TRUE) {
raw_idata = import(here("01_datasets/pumf-98M0001-E-2016-individuals_F1.sav"), 
                      format = "sav", user_na = c("Not available", "88888888", "99999999"))
saveRDS(raw_idata, here("03_output/00_processed_data/idata_raw.Rdata"))} else {
raw_idata =   readRDS(here("03_output/00_processed_data/idata_raw.Rdata"))
}

```

### clean up column names
We want snake-case column names for easier access. ColNames are defined in `ColNamesDict.idata.csv`
```{r}

dict_idata = tibble(old_name = names(raw_idata), explanation = var_label(raw_idata))
#export(dict_idata, file = here("01_datasets/ColNamesDict_idata.csv"))
individ_column_names = import(here("01_datasets/ColNamesDict_idata.csv"))


# we only want to rename columns that we created a new name for
individ_column_names_keep = individ_column_names %>%
  filter(new_name!="")
# rename columns to new name
# pull() is necessary to convert column to vector (which rename_at expects)
raw_idata = raw_idata %>% rename_at(vars(pull(individ_column_names_keep,old_name)), ~ pull(individ_column_names_keep,new_name))



```
### convert to categorical variables to factors to see variable labels
the spss variable has all variables stored as numeric, which makes them read into R as numeric (which means it doesn't show labels present in the dataset). Can use factorize to remedy that 

```{r}
# from the column dictionary, keep only columns that are categorical 
columns_to_convert_to_factor = individ_column_names %>%
  filter(type == "categorical") %>%
  mutate(current_name = if_else(new_name== "", old_name, new_name)) %>% # some of these columns have been renamed, use new name if it was renamed
  pull(current_name)


# batch convert to factor
# this takes a bit long
raw_idata = raw_idata %>%
  mutate_at(.vars = all_of(columns_to_convert_to_factor),
             .funs = factorize)
# now, instead of numbers e.g. for province we see the actual label in the df
```



### select variables of interest
For an easier to manage dataset.
keep all that we have renamed (to include more, edit the `ColNamesDict.csv` in datasets folder)

```{r}

idata = raw_idata %>% select(sort(pull(individ_column_names_keep, new_name)))
```




## clean up age variables
make it easier to differentiate children and adults in the dataset
```{r}


# create age_bracket variable... 
idata$age_bracket = ifelse(as.integer(idata$age_group)>6, "adult","child") # the 6th level is "15 to 17 years"
# check that the result makes sense
table( idata$age_group, idata$age_bracket)

```


## use subset or full data (for troubleshooting)
Usually, this will be set to `use_reduced_data_set == FALSE` unless we are troubleshooting something and want to work with a smaller dataset for faster computation
```{r}
if(use_reduced_data_set == TRUE) {
  idata = idata %>%
    sample_n(50000)
}
```
# cleanup language variables
We need to get from the format of the language questions in the census (see Language Reference Guide) to a categorization of people as bilingual/monolingual, knowing the specific language pairs someone speaks, etc. 
### simplify language factor levels
The response options in the language questions are super long, cut them down to only TRUE/FALSE to be able to reference them more easily
```{r}
# remove the long label names and just keep the true/false label, for easier use later
idata = idata %>%
  mutate_at( .vars = vars(lang_most_eng, lang_most_fre, lang_regular_eng, lang_regular_fre, lang_regular_oth),
             ~ fct_relabel(., ~str_remove(., '\\s*-.*')))


```



### clean up Write-in component for non-official languages (language heard most)
make `lang_most_oth_write_in` more manageable
- Step 1: create a column that is empty if no other language spoken, NA if no info available, and has the name of the language if there is another language spoken
- Step 2: clean up the language names in that variable (make them short & easy to reference)
- Step 3: create a true/false variable for that variable, like we have for English & French
```{r}

idata = idata %>%
  
  mutate(
    # create new column that has the name of the language spoken, with options for no lang/missing info/etc.
    lang_most_oth_name = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "", # when no other languages, put empty string
    lang_most_oth_write_in == "Not available" ~ NA_character_, # when info is missing, put NA
    lang_most_oth_write_in == "All other languages" ~ "Other", # shorten "all other lang" to just "other", for convenience
    
    # now, we try to shorten the write-in component language names
    # if it is a catch-all category, get rid of " languages" to have just "AustroAsiatic", for convenience
    str_detect(lang_most_oth_write_in, "languages")  ~ sub(" languages", "", lang_most_oth_write_in),
     # whenever there is additional info in brackets, remove
    TRUE ~  sub(" \\(.*", "", lang_most_oth_write_in) #HK: this step is removing "(Farsi)", but the next step removes "Persian", so this language is getting erased
  ),
  # remove additional names for languages
  lang_most_oth_name = str_remove(lang_most_oth_name, "Panjabi|PilipinoFilipino") #HK removed "Persian" from here to retain this language

  )

# clean up naming convention to camel case to make it easier to paste and combine later (spaces are removed)
idata$lang_most_oth_name = to_any_case(as.character(idata$lang_most_oth_name), case = "big_camel")


# create a variable that functions like for English and French that just says true/false for lang_most_oth_write_in  
idata = idata %>% 
  mutate(lang_most_oth = case_when(
    lang_most_oth_name == "" ~ "False",
    lang_most_oth_name == "Not available" ~ NA_character_,
    is.na(lang_most_oth_name) ~ NA_character_,
    !is.na(lang_most_oth_write_in) ~ "True"
  ))
```



### calculate number of languages spoken (individual)
#### combining language used most and regularly
That is, if Eng & Fr are used regularly, and German sometimes, n =3
```{r}
# count how many languages each person speaks:
true_false_language_columns = c("lang_most_eng", "lang_most_fre", "lang_most_oth","lang_regular_eng", "lang_regular_fre", "lang_regular_oth")

idata = idata  %>% 
  mutate(n_languages_numeric = rowSums(.[true_false_language_columns] == "True", na.rm = TRUE),
         n_languages = # make this a categorical variable
           case_when(n_languages_numeric == 0 ~ NA_character_,
                     n_languages_numeric == 1 ~ "One Language",
                     n_languages_numeric > 1 ~ "Two or more",
                     TRUE ~ NA_character_))

# to check result, use:
#View(idata %>% select( person_id, lang_most_eng:lang_regular_oth, lang_most_oth, n_languages))

```

#### counting language used most and regularly separately
In case we want to do calculations only on language used most, or regularly used languages
That is, if Eng & Fr are used regularly, and German sometimes, n(languages used most) = 2, n (languages used regularly) = 1
```{r}
#create  a separate vector for the column names that contain info on each
true_false_language_used_most = c("lang_most_eng", 
                                  "lang_most_fre", 
                                  "lang_most_oth")
true_false_language_used_regularly = c("lang_regular_eng", 
                                       "lang_regular_fre", 
                                       "lang_regular_oth")

# count how many languages each person speaks:
idata = idata  %>% 
  mutate(n_languages_used_most = rowSums(.[true_false_language_used_most] == "True", na.rm = TRUE),
         n_languages_used_regularly = rowSums(.[true_false_language_used_regularly] == "True", na.rm = TRUE))
```


### create language pairs variable
We need a column that says what languages that person speaks. So English_French, English&Italian&Other etc. 
If a person speaks a language at home (value is "True"), then copy the column name to new column lang_pair_ordered
```{r}


# sorted by most spoken then regularly spoken languages (for the person)
# concatenate language names with a "&"
idata$lang_pair_most <- apply(idata[true_false_language_used_most], 1, function(x) paste(names(x[x=="True"]), collapse="&"))
idata$lang_pair_regular <- apply(idata[true_false_language_used_regularly], 1, function(x) paste(names(x[x=="True"]), collapse="&"))



```

### clean up language pair variables
remove NAs that are introduced when info is missing, clean up the name of the language from column name to name of language
```{r}
# this takes a long time to run unfortunately
# now we need to clean up the messy column names in each row:
idata = idata %>%
  mutate( # if there is only one language, lang_pair_most is NA 
   lang_pair_most = ifelse(n_languages_numeric <2, NA_character_, lang_pair_most),
       # clean up lang_pair_most values
   lang_pair_most = str_remove(lang_pair_most, 
                               "(NA&)+|(&NA)+"), # remove NA in beginning and middle
   lang_pair_most = str_remove(lang_pair_most, 
                               "(&NA)+|(NA)+"), # remove NAs at End
   lang_pair_most = str_replace(lang_pair_most, 
                                "lang_most_eng", 
                                "English"), # substitute English column names with word english
   lang_pair_most = str_replace(lang_pair_most, 
                                "lang_most_fre", 
                                "French"), # same for french
   lang_pair_most = str_replace(lang_pair_most, 
                                "lang_most_oth", 
                                as.character(lang_most_oth_name)),
  )

# for languages heard regularly
idata = idata %>%
  mutate( # if there is only one language, lang_pair_regular is NA
    lang_pair_regular = ifelse(n_languages_numeric <2, 
                               NA_character_, 
                               lang_pair_regular),
    # clean up lang_pair regular values
    lang_pair_regular = str_remove(lang_pair_regular, 
                                   "(NA&)+|(&NA)+"), # remove NA in beginning and middle
    lang_pair_regular =str_remove(lang_pair_regular,
                                  "&NA"), # remove NAs at End
    lang_pair_regular = str_replace(lang_pair_regular, 
                                    "lang_regular_eng", 
                                    "English"), # substitute English column names with word english
    lang_pair_regular = str_replace(lang_pair_regular, 
                                    "lang_regular_fre", 
                                    "French"), # same for french
    lang_pair_regular = str_replace(lang_pair_regular, 
                                    "lang_regular_oth", 
                                    "Other") # for lang_regular_oth we don't know language, just call it other
   )

```
###combine languages used most and languages used regularly
for now, we don't distinguish between those two!
```{r}

idata$lang_pair =  as.vector(apply(cbind(idata$lang_pair_most, idata$lang_pair_regular), 1, 
             function(x) paste(x[!is.na(x)& !x==""], collapse = "&")))
  idata %>% count(lang_pair) %>% arrange(-n) %>% View()
```

### sort languages from a-z
needed to create language pair variable that does not distinguish between "English & French" and "French & English"
```{r}

# get the individual language names so we can combine them into pairs
idata = idata %>%
  separate(col = lang_pair, into = c("lang1", "lang2", "lang3", "lang4"), sep = "&")

# remove order by most spoken then regularly spoken languages (now sorted by A-Z)
# this takes long to run
idata$lang_pair <- apply(idata[c("lang1", "lang2", "lang3","lang4")], 1, function(x) paste(sort(x[!is.na(x)]), collapse="&"))

# get the sorted languages back
# get the individual language names so we can combine them into pairs
idata = idata %>%
  separate(col = lang_pair, into = c("lang1", "lang2", "lang3", "lang4"), sep = "&", remove = F)

```





# create variable for geographical locations of interest
Make some of the area names more manageable
```{r}

idata$area = fct_recode(idata$area, 
                        Other = "Other census metropolitan areas, census agglomerations and o",
                        `Greater Sudbury – Thunder Bay` = "Greater Sudbury / Grand Sudbury – Thunder Bay")

```


 

#  save data for future analyses
Geo_Percentages and Language pair script use this output
```{r}

idata_child = idata %>%
  filter(age_group %in% c("0 to 4 years", "5 to 6 years", "7 to 9 years"))
saveRDS(idata_child, file = here::here("03_output","00_processed_data", "Individuals_child_data.rdata"))

  

```




#  Individuals dataset: descriptives
### summary tables: children across canada
check how distributions over province and metropolitain areas are
```{r}
# select only adults who are living with at least one child
idata_hh_child = idata %>% 
  filter(n_child_0_to_1 =="One or more" | n_child_2_to_5 =="One or more") %>% 
  filter(age_bracket=="adult")

# how many are there per province & area?
idata_hh_child %>% group_by(province, area) %>%
  summarize(N=n()) %>% 
  spread(province, N)

# breakdown by province
idata_child %>% group_by(province) %>%
  summarize(N=n())

# some provinces have  few data points. can probably concentrate on BC, Alberta, Ontario, Quebec



# top metropolitain areas
idata_child %>% group_by(area) %>%
   summarize(N=n()) %>%
  arrange(-N)
# top metropolitain areas are: Toronto, Montréal, Vancouver Calgary, Ottawa - Gatineau, Edmonton (cutoff could be 5000 individuals?)

```
## bilingualism in children in individuals dataset
#### summary table: bilingual children across canada

```{r}
 (canada_percent = idata %>% 
  mutate(country = "Canada",
    age_collapsed = fct_collapse(age_group, 
                      "age_0_to_4" = "0 to 4 years",
                      "age_5_to_9" = c("5 to 6 years", "7 to 9 years"))) %>%
  filter(age_collapsed %in% c("age_0_to_4", "age_5_to_9")) %>%
   group_by(country, age_collapsed, n_languages) %>%
   summarize(N=n()) %>%
   spread(n_languages, N) %>%
   clean_names() %>%
   mutate(total = one_language+two_or_more, 
          percent = round(two_or_more/total,5)*100)) #HK note: here the percent doesn't use NA in the denominator, but below on line 395, it does. Is this on purpose?


# canada 0-9 
canada_percent %>% group_by(country) %>%
  summarize(percent = sum(two_or_more)/sum(total+na))
```





# save data for future analysis
```{r}
export(idata, file = here::here("03_output","00_processed_data", "idata_preprocessed.Rdata")) 
beepr::beep("ping")
```

