---
title: "Bilingualism in Canada: Geographic Distribution Tables"
author: Esther Schott (esther.schott@mail.concordia.ca)
output: html_document
---
output: 
  html_document: 
    code_folding: show
    collapsed: no
    df_print: kable
    highlight: espresso
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\\\
# load libraries
```{r load packages}
#install.packages("gpclib", type="source")

library(tidyverse)
library(rgeos)
library(rgdal)
library(maptools)
library(sp)
library(here)
library(fuzzyjoin)
library(viridis)
library(scales)
library(ggrepel)
library(rio)
```





# process census language data
## load census language data
```{r read_census_data}
idata_child = import(file = here::here("03_output","00_processed_data", "idata_preprocessed.Rdata")) %>%
  filter(age_group %in% "0 to 4 years")

```

## summarize census language data by area & province
```{r summarize_census_data}
(gma_LangGroup_data = idata_child %>%
  group_by(province,area,n_languages) %>%
  summarize(N=n()) %>%
    mutate( Percent = round(N/sum(N),2)) %>% 
    filter(n_languages == "Two or more") ) %>%
  ungroup
    

```
# table for poster
```{r}
table_prov = idata_child %>%
  group_by(province) %>%
  mutate(total = n()) %>% 
  group_by(province,total,n_languages) %>%
  summarize(N=n()) %>% 
    mutate( Percent = scales::percent(N/total,1))%>% 
    filter(n_languages != "One Language")  %>%
  select(-n_languages)

table_city = idata_child %>%
  group_by(province,area,n_languages) %>%
  summarize(N=n()) %>%
  
    mutate( Percent_city = scales::percent(N/sum(N),1)) %>%
  ungroup() %>%
mutate(      area = ifelse(area =="Other","_other",as.character(area)))%>% 
    filter(n_languages != "One Language")  %>%
  select(-n_languages) %>%
  arrange(area,-N) %>% filter(N >250 ) 


```



## process "other" non-cma area data
```{r}

province_data = gma_LangGroup_data %>%    filter(substr(area,1,5) == "Other")

province_data$province= as.character(province_data$province)

northern_data = data.frame(province= rep("Northern Canada",3), id = c("Yukon", "Northwest Territories", "Nunavut"))

province_data  = province_data %>% full_join(northern_data) %>%
  mutate(id = ifelse(is.na(id), as.character(province), as.character(id)))

```
## save percentages for maps
```{r}
export(province_data, here("03_output", "00_processed_data", "idata_geo_province.Rdata"))
gma_LangGroup_data %>% 
  filter(area != "Other") %>%
export(., here("03_output", "00_processed_data", "idata_geo_area.Rdata"))

```



## save data for poster table
```{r}
table_otherCMA = province_data %>% group_by(province) %>% slice(1) %>% ungroup() %>% select(-id, -N) %>%
  mutate(Percent = scales::percent(Percent, 1))


table_poster = table_prov %>% full_join(table_city %>% 
  rename( Percent = Percent_city))  %>% 
  full_join(table_otherCMA) %>%
  select(province, area,total, N, Percent) %>%
  arrange(province)
write.csv(table_poster, file=here("03_output","02_tables","Table_idata_by_area.csv"),row.names = F)
```




