---
title: "Census Bilingualism Language Pair Analysis"
output:
  html_document:
    df_print: paged
---
author: Esther Schott
please email esther.schott@mail.concordia.ca if you have any questions
see poster created from earlier versions this script: https://osf.io/4gekn/

# 1. Preparation Steps
## Script settings
```{r}

```

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(rio) # for more powerful import & export of data
library(labelled)
library(janitor) # cleaning up variable names
library(codebook) # better documentation
library(snakecase)

```

## read in data
```{r}
idata = import(here::here("03_output","00_processed_data", "idata_preprocessed.Rdata"))
```


# create a long data frame with all language pairings per person
The goal of this is to show which languages are often combined with each other. For example, how many ppl are in Montreal speaking Arabic and French vs in Edmonton. It's okay if these people also speak another language. Therefore, we will create an extra dataframe that has multiple rows for trilingual ppl 
```{r}

# create all possible pairings
idata = idata %>% 
  mutate(pairing_1 = ifelse(!is.na(lang1) & !is.na(lang2), 
                             paste(lang1, lang2, sep = "&"), NA_character_),
         pairing_2 =        ifelse(!is.na(lang1) & !is.na(lang3), 
                             paste(lang1, lang3, sep = "&"), NA_character_),
         pairing_3  =       ifelse(!is.na(lang2) & !is.na(lang3), 
                             paste(lang2, lang3, sep = "&"), NA_character_),
         pairing_4 =        ifelse(!is.na(lang1) & !is.na(lang4), 
                             paste(lang1, lang4, sep = "&"), NA_character_),
         pairing_5 =        ifelse(!is.na(lang2) & !is.na(lang4), 
                             paste(lang2, lang4, sep = "&"), NA_character_),
         pairing_6 =        ifelse(!is.na(lang3) & !is.na(lang4), 
                              paste(lang3, lang4, sep = "&"), NA_character_))
                            
lang_pair_data = idata %>%
  filter(n_languages_numeric>1) %>%
  gather(number, language_pair, pairing_1:pairing_6) %>%
  filter(!is.na(language_pair))
  

```

# languages in canada overall
```{r}
lang_pair_data %>% filter(age_group == "0 to 4 years") %>% group_by(age_group) %>% count(language_pair) %>% mutate(percent = round(n/sum(n)*100,1)) %>% arrange(-percent) %>% top_n(10) %>% View()
```

## fixing language combo names
```{r}
language_pair_dictionary = tibble(language_pair_old = unique(lang_pair_data$language_pair)) %>%
  filter(str_detect(language_pair_old, "Other|AustroAsiatic", negate = TRUE)) %>%
  mutate(language_pair = gsub("&", " & ", language_pair_old),
         fix = str_detect( language_pair, "^English|^French|^Abo", negate = TRUE)) %>% 
  separate(language_pair, into = c("lang1", "lang2"), sep = " & ", remove = FALSE) %>%
  mutate(language_pair = ifelse(fix == TRUE, paste(lang2, lang1, sep = " & "), language_pair)) %>%
  select(language_pair, language_pair_old)



```


# Table 2: look at languages by province
## prep 
```{r}
age_groups_of_interest = c("0 to 4 years", "5 to 6 years", "7 to 9 years")

# how many children are in the census dataset (regardless of languages spoken)
n_children_by_age = idata %>% filter(age_group %in% age_groups_of_interest) %>%
  mutate(age_collapsed = fct_collapse(age_group, 
                      "age_0_to_4" = "0 to 4 years",
                      "age_5_to_9" = c("5 to 6 years", "7 to 9 years"))) %>% 
  group_by(province, age_collapsed) %>% 
  count(name = "total_by_age") %>%
  ungroup()



```

## percentage out of bilingual/all children by age group
 separate for 0-4 and 5-9 yo
```{r}

lang_pair_by_province_and_age_long  = 
  lang_pair_data %>% 
  filter(age_group %in% age_groups_of_interest) %>%
  # collapse two older ages into one category
  mutate(age_collapsed = fct_collapse(age_group, 
                      "age_0_to_4" = "0 to 4 years",
                      "age_5_to_9" = c("5 to 6 years", "7 to 9 years"))) %>% 
  group_by(province, age_collapsed) %>% 
  count(language_pair) %>% 
  left_join(n_children_by_age) %>% # how how many children per age group
  mutate(n_bilingual_children = sum(n),
    percent_bilingual_children = round(n/n_bilingual_children*100,1), # add percent
    percent_all_children = round(n/total_by_age*100,1)) %>% 
  ungroup() 

lang_pair_by_province_and_age = lang_pair_by_province_and_age_long %>%
  select( - n, -total_by_age, - n_bilingual_children) %>% 
  pivot_wider(names_from = age_collapsed, values_from = c("percent_bilingual_children","percent_all_children"))

```
### canada-wide
```{r}
totals_across_canada = lang_pair_by_province_and_age_long %>% 
  mutate(country = "Canada") %>% 
  group_by(province, age_collapsed) %>%
  top_n(n, n = 1) %>%
  select(country, province, age_collapsed,  n_bilingual_children, total_by_age)%>%
  group_by(country, age_collapsed) %>%
  summarize(total_by_age_canada = sum(total_by_age),
            n_bilingual_by_age_canada = sum(n_bilingual_children))

canada_language_pairs = lang_pair_by_province_and_age_long %>% 
  left_join(totals_across_canada) %>% 
 group_by(country, age_collapsed, language_pair, total_by_age_canada, n_bilingual_by_age_canada) %>%
      summarise(n_canada = sum(n)) %>%
  mutate(
               percent_all_children =  round(n_canada/total_by_age_canada*100,5),
               percent_bilingual_children = round(n_canada/n_bilingual_by_age_canada*100,5)) %>% 
  arrange(age_collapsed, -percent_bilingual_children) %>%
  ungroup()

canada_language_pairs_wide = canada_language_pairs %>% select( - n_canada, -total_by_age_canada, - n_bilingual_by_age_canada) %>% 
  pivot_wider(names_from = age_collapsed, 
              values_from = c("percent_bilingual_children", "percent_all_children"))

     
```


## percentage out of all children, combined for ages 0-9
### canada-wide
```{r}
canada_language_pairs_totals = 
  canada_language_pairs %>% 
  distinct(age_collapsed, .keep_all = TRUE) %>% 
  select(country, age_collapsed, total_by_age_canada, n_bilingual_by_age_canada) %>% 
  summarize(total_by_age_canada = sum(total_by_age_canada),
            n_bilingual_by_age_canada = sum(n_bilingual_by_age_canada)) %>%
  mutate(age_collapsed = "age_0_to_9")

canada_language_pair_data = canada_language_pairs %>%
   mutate(age_collapsed = fct_collapse(age_collapsed, 
                            "age_0_to_9" = c("age_0_to_4", "age_5_to_9"))) %>%
  group_by(language_pair, age_collapsed) %>%
  summarize(n = sum(n_canada)) %>%
  full_join(canada_language_pairs_totals) %>%
  mutate(
               percent_all_children =  round(n/total_by_age_canada*100,5),
               percent_bilingual_children = round(n/n_bilingual_by_age_canada*100,5)) %>% 
  arrange(age_collapsed, -percent_bilingual_children) %>%
  ungroup() %>%
  select( - n, -total_by_age_canada, - n_bilingual_by_age_canada) %>% 
  pivot_wider(names_from = age_collapsed, 
              values_from = c("percent_bilingual_children", "percent_all_children"))

```
### merge both by age and 0-9 together
```{r}
canada_language_pair_data = canada_language_pair_data %>%
  full_join(canada_language_pairs_wide)
```

### provinces
```{r}
lang_pair_by_province_all_ages_combined = lang_pair_data %>% 
  
  # collapse two older ages into one category
  mutate(age_combined = fct_collapse(age_group, 
                            "age_0_to_9" = age_groups_of_interest)) %>%
  filter(age_combined == "age_0_to_9") %>%
  # count how many children per language pair
  group_by(province) %>%
  count(language_pair) %>% 
  # merge in number of children collapsed across ages
   full_join(n_children_by_age %>% group_by(province) %>% summarize(total_by_age = sum(total_by_age))) %>%
  #calculate percentages
  
   mutate(percent_bilingual_children_age_0_to_9 = round(n/sum(n)*100,1), # add percent
         percent_all_children_age_0_to_9 = round(n/total_by_age*100, 1)) %>% 
  select(-n, -total_by_age)
  
```



## combine detailed age breakdown and summary across all ages of interest
```{r}
# 
lang_pair_by_province = lang_pair_by_province_and_age %>% 
  full_join(lang_pair_by_province_all_ages_combined) %>% 
  # add canada data
  full_join(canada_language_pair_data) %>%
  # remove other label
  mutate( other_flag = str_detect(language_pair, "Other")) %>%
  filter(other_flag != TRUE) %>% 
  # exclude language combos that appear very rarely
  filter(percent_bilingual_children_age_0_to_9 >= 4|
         country == "Canada" & percent_bilingual_children_age_0_to_9 >= 3) %>% 
  select(-other_flag) %>%
  arrange(province, -percent_bilingual_children_age_0_to_9) %>%
    # fix language pair names
  rename(language_pair_old = language_pair) %>% 
  left_join(language_pair_dictionary, by = "language_pair_old") %>%
  select(-language_pair_old) %>%
  relocate(language_pair, 
           .after = province) %>%
  relocate(country, .before = province) %>%
  relocate(percent_bilingual_children_age_0_to_9, 
           .after = percent_bilingual_children_age_5_to_9)


lang_pair_by_province %>%
  write_csv(here("03_output", "02_tables", "language_pairs_by_province.csv"))




```






# save data for future analysis
```{r}
export(idata, file = here::here("03_output","00_processed_data", "idata_preprocessed.Rdata")) 
beepr::beep("ping")
```

