---
title: "Census Bilingualism Pre-processing HOUSEHOLDS"
output:
  html_document:
    df_print: paged
---
author: Esther Schott
please email esther.schott@mail.concordia.ca if you have any questions
see poster created from a previous version of this script: https://osf.io/4gekn/

# 1. Preparation Steps
## Script settings
```{r}
load_data_from_raw = FALSE # If true, loads data from .sav, else loads R file (faster)

use_reduced_data_set = TRUE # If true, uses a smaller subset of full data set (for troubleshooting etc.)
```

## load packages
```{r echo=FALSE}
library(here)
library(tidyverse)
library(rio) # for reading in SPSS files
library(labelled)
#library(cowplot)
#library(viridis)
library(janitor)

library(codebook)


library(scales)

library(lme4)

```



#  HOUSEHOLDS dataset




## read in dataset 
```{r}
# need to change to reading in factor solution as above
if(load_data_from_raw == TRUE) {
  raw_hhdata = rio::import(here("01_datasets/pumf-98M0002-E-2016-hierarchical_F1.sav"),format = "sav", user_na = c("Not available", "88888888", "99999999"))
  saveRDS(raw_hhdata, here("03_output/00_processed_data/hdata_raw.Rdata"))} else {
  raw_hhdata = readRDS(here("03_output/00_processed_data/hdata_raw.Rdata"))
}

#dict = tibble(old_name = names(raw_hhdata), explanation = var_label(raw_hhdata))
#export(dict, file = here("01_datasets/ColNamesDict.csv"))
hh_column_names = import(here("01_datasets/ColNamesDict_hhdata.csv"))




```
## data cleaning - general

### clean up column names
```{r}
# we only want to rename columns that we created a new name for
hh_column_names_keep = hh_column_names %>%
  filter(new_name!="")
# rename columns to new name
# pull() is necessary to convert column to vector (which rename_at expects)
raw_hhdata = raw_hhdata %>% rename_at(vars(pull(hh_column_names_keep,old_name)), ~ pull(hh_column_names_keep,new_name))
```
### convert to categorical variables to factors to see variable labels
```{r}
# make factor - the spss variable has all variables stored as numeric, which makes them read into R as numeric (which means it doesn't show labels present in the dataset). Can use factorize to remedy that 

# from the column dictionary, keep only columns that are categorical 
columns_to_convert_to_factor = hh_column_names %>%
  filter(type == "categorical") %>%
  mutate(current_name = if_else(new_name== "", old_name, new_name)) %>% # some of these columns have been renamed, use new name if it was renamed
  pull(current_name)


# batch convert to factor
# this takes a bit long
raw_hhdata = raw_hhdata %>%
  mutate_at(.vars = vars(columns_to_convert_to_factor),
             .funs = factorize)
# now, instead of numbers e.g. for provice we see the actual label in the df
```



### select variables of interest
```{r}
# select which variables are necessary for our analysis
# keep all that we have renamed (to include more, edit the ColNamesDict.csv in datasets folder)

hhdata = raw_hhdata %>% select(pull(hh_column_names_keep, new_name))
```





### create variable for households with children 9 and under

```{r}
# create a variable that says whether child younger than 10 lives in household


hhdata = hhdata %>% 
  mutate(child0_9 = ifelse(age_group == "0 to 9 years", 1,0)) %>%
    group_by(household_id) %>%
  mutate(children_per_hh = sum(child0_9)) %>%
  ungroup()



```

### use subset of full data (for troubleshooting)
This is necessary while creating the code because some code is SLOW. change use_reduced_data_set to change at the top of the script
```{r}

if(use_reduced_data_set == TRUE) {
  # select X unique households to keep in subset
  household_subsample = sample(unique(hhdata$household_id),
                               20000)  
  hhdata = hhdata %>%
    filter(household_id %in% household_subsample)
}

```

### clean up area variable
```{r}
hhdata = hhdata %>%
  mutate(area = sub(" .*", "", area))

```



## data cleaning -   language variables
```{r}

#make this simpler


hhdata = hhdata %>%
  mutate_at( .vars = vars(lang_most_eng, lang_most_fre, lang_regular_eng, lang_regular_fre, lang_regular_oth),
             ~ fct_relabel(., ~str_remove(., '\\s*-.*')))

# create a variable that functions like all other home language variables and just says true/false for lang_most_oth_write_in  
hhdata = hhdata %>% 
  mutate(lang_most_oth = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "False",
    lang_most_oth_write_in == "Not available" ~ NA_character_,
    !is.na(lang_most_oth_write_in) ~ "True"
  ))
# test that this is working: 
# table(hhdata$lang_most_oth_write_in, hhdata$lang_most_oth)

###LK: do the same for mother tongue?

```
### clean up Write-in component for non-official languages
```{r}
hhdata = hhdata %>%
  
  mutate(lang_most_oth_write_in = case_when(
    lang_most_oth_write_in == "No non-official language" ~ "",
    lang_most_oth_write_in == "Not available" ~ NA_character_,
    lang_most_oth_write_in == "All other single languages" ~ "Other",
    TRUE ~  sub(" .*", "", lang_most_oth_write_in)
  ))
    
```



### calculate number of languages spoken (individual)

```{r}
# count how many lanuages each person speaks:
true_false_language_columns = c("lang_most_eng", "lang_most_fre", "lang_most_oth","lang_regular_eng", "lang_regular_fre", "lang_regular_oth")

hhdata = hhdata  %>% 
  mutate(N_languages_person_numeric = rowSums(.[true_false_language_columns] == "True", na.rm = TRUE),
         N_languages_person = # make this a categorical variable
           case_when(N_languages_person_numeric == 0 ~ NA_character_,
                     N_languages_person_numeric == 1 ~ "One Language",
                     N_languages_person_numeric > 1 ~ "Two or more",
                     TRUE ~ NA_character_))

# check result:
#View(hhdata %>% select(household_id, person_id, lang_most_eng:lang_regular_oth, lang_most_oth, N_languages_individual))


```

### create language pairs variable

```{r}
# we need a column that says what languages that person speaks. So English_French, English&Italian&Other etc. 
# if a person speaks a language at home (value is "True"), then copy the column name to new column lang_pair_ordered

# sorted by most spoken then regularly spoken languages (for the person)
hhdata$lang_pair_ordered <- apply(hhdata[true_false_language_columns], 1, function(x) paste(names(x[x=="True"]), collapse="&"))



# this takes a long time to run 
# now we need to clean up the messy column names in each row:
hhdata = hhdata %>%
  mutate( lang_pair_ordered = str_remove(lang_pair_ordered, "(NA&)+|(&NA)+"), # remove NA in beginning and middle
   lang_pair_ordered =str_remove(lang_pair_ordered,"&NA"), # remove NAs at End
   lang_pair_ordered = str_replace(lang_pair_ordered, "(lang_most_eng)+|(lang_regular_eng)+", "English"), # substite English column names with word english
   lang_pair_ordered = str_replace(lang_pair_ordered, "(lang_most_fre)+|(lang_regular_fre)+", "French"), # same for french
   lang_pair_ordered = str_replace(lang_pair_ordered, "lang_most_oth", as.character(lang_most_oth_write_in)),
   lang_pair_ordered = str_replace(lang_pair_ordered, "lang_regular_oth", "Other") # for lang_regular_oth we don't know language, just call it other
   )

# check result:
#View(hhdata %>% select(household_id, person_id, lang_most_eng:lang_regular_oth, lang_most_oth, N_languages_individual, lang_pair_ordered))




```
### create variables for each language in dataset
```{r}

hhdata  =  hhdata %>%
  separate(lang_pair_ordered, c("lang1", "lang2", "lang3", "lang4"), remove = FALSE)
# warning that not all people have 4 languages is normal
```



### simplify language pairs variable (sorted A-Z)
```{r}

# remove order by most spoken then regularly spoken languages (now sorted by A-Z)
hhdata$lang_pair <- apply(hhdata[c("lang1", "lang2", "lang3","lang4")], 1, function(x) paste(sort(x[!is.na(x)]), collapse="&"))

```
## get columns back in alphabetical order
```{r}
hhdata = hhdata %>% select(sort(names(hhdata)))
```


# Descriptives
## keep only households with kids under 10
```{r}
# restrict to kids under 10
# keep only households that have at least one child under 10
hhdata_child = hhdata %>%
  filter(children_per_hh >0)
```


## summary tables: children under 9 years
```{r}


# number of bilinguals
#canadian average:
hhdata_child %>%
filter(age_group=="0 to 9 years") %>%
  group_by(N_languages_person) %>%
  summarize(N=n()) %>%
    mutate( Percent = N/sum(N)) 





(area_lang_data = hhdata_child %>%
filter(age_group=="0 to 9 years") %>%
  group_by(province,area,N_languages_person) %>%
  summarize(N=n()) %>%
    mutate( Percent = round(N/sum(N),3))%>%
    filter(N_languages_person != "One Language"))


# # specific language combos
 hhdata_child_langPair = hhdata_child %>%
   filter(age_group=="0 to 9 years",N_languages_person_numeric>1 ) %>%
   group_by(province,lang_pair) %>%
   summarize(N=n())%>%
   mutate( Percent = round(N/sum(N),3))%>%
   arrange(province, -N)

 hhdata_child_langPair=hhdata_child_langPair %>% group_by(province) %>%
   slice(1:5) %>%
   mutate(id=seq_along(Percent))

# add nicer language labels to graph (need to adapt to new data)
#lang_dict = read_csv(file= here("01_datasets","LangPair_Dict_2016.csv"))
#hhdata_child_langPair = full_join(hhdata_child_langPair, lang_dict, by = c("lang_pair"= "Lang_Combo_top2" )


```


# export for analysis in other scripts
```{r}

# this contains ALL households (in case another analysis needs all data)
export(hhdata, file = here("03_output/00_processed_data/hhdata_preprocessed.Rdata"))
```





# getting data from only adults in child's households
```{r}
# filter out children
hhdata_child_adult <- hhdata_child %>% 
  filter(child0_9 == 0)
  
# see number of adults per household
adults_n <- hhdata_child_adult %>% 
  # group by household
  group_by(household_id) %>% 
  summarize(n_adults = n())
# number of adults per household ranges 1 - 6



### language

# get language (pairs) and number of speakers per household
adults_lang <- hhdata_child_adult %>% 
  group_by(household_id, lang_pair) %>% 
  summarize(n_lang_pair = n()) 

# get number of bilingual adults per household
adults_biling_n <- adults_lang %>% 
  filter(str_detect(lang_pair, "&")) %>% 
  group_by(household_id) %>% 
  summarize(n_biling_adults = sum(n_lang_pair))

# get proportion of bilingual adults per household
adults_n <- adults_n %>% 
  left_join(adults_biling_n, by = "household_id") %>% 
  mutate(n_biling_adults = replace_na(n_biling_adults, 0), 
         prop_biling_adults = n_biling_adults / n_adults)



### education

# currently 8 factor levels. want to change to 5 to match Turcotte analyses. merging post-secondary levels together

hhdata_child_adult <- hhdata_child_adult %>% 
  mutate(education_rank = case_when(
    education == "No certificate, diploma or degree" ~ 1,
    education == "Secondary (high) school diploma or equivalency certificate" |
    education ==  "Trades certificate or diploma other than Certificate of Appr" |
    education ==  "Certificate of Apprenticeship or Certificate of Qualificatio" |
    education ==  "College, CEGEP or other non-university certificate or diplom" ~ 2,
    education ==  "University certificate or diploma below bachelor level"  ~ 3,
    education == "Bachelor's degree" ~ 4, 
    education == "University certificate, diploma or degree above bachelor lev" ~ 5,
    education == "Not applicable" ~ 0
  ))

adults_edu <- hhdata_child_adult %>% 
  group_by(household_id) %>% 
  summarize(highest_ed = max(education_rank)) %>% 
  mutate(highest_ed = as.factor(highest_ed))

```


# create df for only children

```{r}
hhdata_child_only <- hhdata_child %>% 
  # filter for only children
  filter(child0_9 == 1) %>% 
  # join with # of bilingual adults
  left_join(adults_n, by = "household_id") %>% 
  # join with highest education
  left_join(adults_edu, by = "household_id")
```


# exploratory data viz

## bilingal adults in household
```{r}
# number of bilingual children by number of bilingual adults in HH (HH without biling adults removed, so more detail can be seen)
hhdata_child_only %>% 
  filter(n_biling_adults != 0) %>% 
ggplot(aes(x = factor(n_biling_adults), fill = N_languages_person)) +
  geom_bar(position = "dodge")


# proportion of bilingual children in HH with various numbers of biling adults
ggplot(hhdata_child_only, aes(x = factor(n_biling_adults), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  geom_abline(intercept = 0.5, slope = 0)



# number of bilingual children in HH with various proportion of bilingual adults
ggplot(hhdata_child_only, aes(x = prop_biling_adults, fill = N_languages_person)) +
  geom_histogram(binwidth = .1)

## same but filtered for no bilingual adults
hhdata_child_only %>% 
  filter(n_biling_adults != 0) %>% 
  ggplot(aes(x = prop_biling_adults, fill = N_languages_person)) +
  geom_histogram(binwidth = .1)




## get total number of children in the database
n_children_total <- nrow(hhdata_child_only)

# percentage of bilingual children based on number of bilingual adults in their HH (O, 1, 2+)

biling_by_adults <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  mutate(n_biling_adults_group = case_when(
    n_biling_adults == 0 ~ "None",
    n_biling_adults == 1 ~ "One Adult",
    n_biling_adults > 1 ~ "Two or More Adults",
    TRUE ~ "NA"
  )) %>% 
  mutate(n_biling_adults_group = as.factor(n_biling_adults_group)) %>% 
  group_by(n_biling_adults_group) %>% 
  summarize(n_children = n(), percent_children = n_children/n_children_total * 100)

```

## province / cities

```{r}
# proportion of bilingual children by province
ggplot(hhdata_child_only, aes(x = province, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))


ggplot(hhdata_child_only, aes(x = province, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90)) + 
  facet_grid(. ~ n_biling_adults)



# proportion of bilingual children in major cities
ggplot(hhdata_child_only, aes(x = area, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))


## get total number of children in each province
n_children_province <- hhdata_child_only %>% 
  group_by(province) %>% 
  summarize(n_children_province = n())

# percentage of bilingual children per province based on number of bilingual adults in their HH (O, 1, 2+)

biling_by_adults_province <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  mutate(n_biling_adults_group = case_when(
    n_biling_adults == 0 ~ "None",
    n_biling_adults == 1 ~ "One Adult",
    n_biling_adults > 1 ~ "Two or More Adults",
    TRUE ~ "NA"
  )) %>% 
  mutate(n_biling_adults_group = as.factor(n_biling_adults_group)) %>% 
  group_by(n_biling_adults_group, province) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_province, by = "province") %>%
  mutate(percent_children = n_children/n_children_province * 100)

```

## parental ed

```{r}
ggplot(hhdata_child_only, aes(x = factor(highest_ed), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))

# by province
ggplot(hhdata_child_only, aes(x = factor(province), fill = N_languages_person)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90))+
  facet_grid( . ~ highest_ed)


## get total number of children for each parental education level
n_children_edu <- hhdata_child_only %>% 
  group_by(highest_ed) %>% 
  summarize(n_children_edu = n())

# percentage of bilingual children based on highest education of parents

biling_by_adults_edu <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(highest_ed) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_edu, by = "highest_ed") %>%
  mutate(percent_children = n_children/n_children_edu * 100)

## get total number of children for each parental education level in each province
n_children_edu_province <- hhdata_child_only %>% 
  group_by(highest_ed, province) %>% 
  summarize(n_children_edu_province = n())

# percentage of bilingual children based on highest education of parents by province
biling_by_adults_edu_province <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(highest_ed, province) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_edu_province, by = c("highest_ed", "province")) %>%
  mutate(percent_children = n_children/n_children_edu_province * 100)

```

## immigration status

```{r}
ggplot(hhdata_child_only, aes(x = immigration_generation, fill = N_languages_person)) +
  geom_bar(position = "fill") 
#+  theme(axis.text.x = element_text(angle = 90))


ggplot(hhdata_child_only, aes(x = immigration_generation, fill = N_languages_person)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ province)


## get total number of children in each immigration generation
n_children_immigration <- hhdata_child_only %>% 
  group_by(immigration_generation) %>% 
  summarize(n_children_immigration = n())

# percentage of bilingual children per immigration generation

biling_by_adults_immigration <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(immigration_generation) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_immigration, by = "immigration_generation") %>%
  mutate(percent_children = n_children/n_children_immigration * 100)


## get total number of children for each immigration generation in each province
n_children_immigration_province <- hhdata_child_only %>% 
  group_by(immigration_generation, province) %>% 
  summarize(n_children_immigration_province = n())

# percentage of bilingual children based on highest education of parents by province
biling_by_adults_immigration_province <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(immigration_generation, province) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_immigration_province, by = c("immigration_generation", "province")) %>%
  mutate(percent_children = n_children/n_children_immigration_province * 100)

```

## gender

```{r}
ggplot(hhdata_child_only, aes(x = sex, fill = N_languages_person)) +
  geom_bar(position = "fill") 
```

## income

```{r}
ggplot(hhdata_child_only, aes(x = income_decile, fill = N_languages_person)) +
  geom_bar(position = "fill") 

## get total number of children in each income decile
n_children_income <- hhdata_child_only %>% 
  group_by(income_decile) %>% 
  summarize(n_children_income = n())

# percentage of bilingual children per immigration generation

biling_by_adults_income <- hhdata_child_only %>% 
  filter(N_languages_person_numeric > 1) %>% 
  group_by(income_decile) %>% 
  summarize(n_children = n()) %>% 
  left_join(n_children_income, by = "income_decile") %>%
  mutate(percent_children = n_children/n_children_income * 100)
```



# model


```{r}

# add column for bilingual hit for regression models
hhdata_child_only <-  hhdata_child_only %>% 
  mutate(bilingual = case_when(
    N_languages_person_numeric == 1 ~ 0,
    N_languages_person_numeric > 1 ~ 1
  )) %>% 
  mutate(bilingual = as.factor(bilingual)) %>% 
  mutate(n_biling_adults_group = case_when(
    n_biling_adults == 0 ~ "None",
    n_biling_adults == 1 ~ "One Adult",
    n_biling_adults > 1 ~ "Two or More Adults",
    TRUE ~ "NA"
  )) 

## need to decide which measure of adult bilingualism to use (N, proportion, group)
m1 <- glm(bilingual ~ n_biling_adults + province + highest_ed + immigration_generation + income_decile, data = hhdata_child_only, family = "binomial")
summary(m1)
```



