---
title: "Bilingualism in Canada: Geographic Distribution Tables"
author: Esther Schott (esther.schott@mail.concordia.ca)
output: html_document
---
output: 
  html_document: 
    code_folding: show
    collapsed: no
    df_print: kable
    highlight: espresso
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\\\
# load libraries
```{r load packages}
#install.packages("gpclib", type="source")

library(tidyverse)
library(rgeos)
library(rgdal)
library(maptools)
library(sp)
library(here)
library(fuzzyjoin)
library(viridis)
library(scales)
library(ggrepel)
library(rio)
library(tidylog)
```





# process census language data
## load census language data
```{r read_census_data}
idata = import(file = here::here("03_output","00_processed_data", "idata_preprocessed.Rdata"))
age_groups_of_interest = c("0 to 4 years", "5 to 6 years", "7 to 9 years")



```

# Table 1: detailed age breakdown
## narrow down age range to kids, select columns
```{r}
idata_child = idata %>%
  # look only at children under 9
  filter(age_group %in% age_groups_of_interest) %>%
  # collapse older age ranges into one
  mutate(age_collapsed = fct_collapse(age_group, 
                      "age_0_to_4" = "0 to 4 years",
                      "age_5_to_9" = c("5 to 6 years", "7 to 9 years"))) %>%
  select(province, area, age_group, age_collapsed, n_languages, lang_pair, sex, person_id)

```
## remove missing data from dataframe
For some children, there was no info on languages spoken, we remove those to avoid underestimating the % of bilinguals
```{r}
idata_child = idata_child %>%
  filter(!is.na(n_languages))
```



## calculate totals, number of bilinguals, and %
All of the following three sections (province, canada-wide, metropolitan area) work fundamentally the same, for three age groups (0-4, 5-9, 0-9) to create the final table for the paper.  
### province-wide
#### 0-4 and 5-9

xxx decide what to do with missing values
```{r}


# calculate province-wide breakdown
table_prov = idata_child %>%
  group_by(province, age_collapsed) %>%
  # get totals for province & age
  mutate(total_for_area_and_age = n()) %>% 
  # group by # of languages too
  group_by(province, age_collapsed, total_for_area_and_age, n_languages) %>%
  # count numbers
  summarize(N=n()) %>%
  # calculate percent
  mutate( Percent = N/total_for_area_and_age)%>% 
  # actually we only care about Two or more languages
  filter(n_languages != "One Language")  %>%
  select(-n_languages) %>%
  rename(N_bilingual = N) %>%
  # necessary for merging with area-level data
  mutate(area = "") %>%
  ungroup()

# convert to wide data for convenience
# shows numbers and % next to each other
table_prov_wide = table_prov %>% 
  select(-total_for_area_and_age) %>%
  pivot_wider(names_from = age_collapsed, 
              values_from = c(Percent, N_bilingual))
```
#### age 0-9
```{r}
table_prov_0_9 = table_prov %>% 
  group_by(province) %>%
  summarise(Percent_age_0_to_9 = sum(N_bilingual)/sum(total_for_area_and_age))
 
```
#### combine separate age groups and age 0-9
```{r}

table_prov_combined = table_prov_wide %>%
  # don't need the actual numbers here, just %
  select(-starts_with("N_biling")) %>%
  full_join(table_prov_0_9)
```



### canada-wide
#### by age subgroup & combined
```{r}

# calculate bilingualism % for all of Canada, to merge into table later
total_across_canada = table_prov %>%
  # combine numbers from all provinces into one
  group_by( age_collapsed) %>%
  # calculate number of children & number of bilingual children
  summarize(total= sum(total_for_area_and_age),
            N_bilingual = sum(N_bilingual)) %>% # need this intermediate step to be able to calculate combined column 0-9
  # calculate percent, add column to signfiy this is canada-wide data
  mutate(Percent = N_bilingual/total,
         country = "Canada") %>% 
  # create wide dataframe for paper
  pivot_wider(names_from = "age_collapsed", values_from = c("total", "N_bilingual", "Percent")) %>% # to be able to add a column on 0 to 9
  # calculate percent for 0-9
  mutate(Percent_age_0_to_9 = sum(N_bilingual_age_0_to_4, N_bilingual_age_5_to_9)/
          sum(total_age_0_to_4, total_age_5_to_9) ) # combined across ages
```
#### keep just relevant columns for paper
```{r}
table_canada = total_across_canada %>% select(-starts_with("total"), -starts_with("N_bili")) 
```


### by metropolitan area 
#### 0-4 and 5-9
```{r}
table_city_by_age_subgroup = idata_child %>% 
  # remove provinces that don't have metropolitan areas in census
  filter(!province %in% c("Newfoundland and Labrador", "Prince Edward Island", "Northern Canada")) %>%
  group_by(province, age_collapsed, area) %>%
  mutate(total = n()) %>% 
  group_by(province, age_collapsed, total, area, n_languages) %>%
  summarize(N=n()) %>% 
  mutate( Percent = round(N/total,3)) %>%
  ungroup() %>% 
  mutate(area = ifelse(area =="Other","zz_other",as.character(area)))%>% # so that other sorts at the end always
  filter(n_languages != "One Language")  %>% 
  rename(N_bilingual = N) %>%
  select(province, area, age_collapsed, total, N_bilingual, Percent) %>%
  arrange(province,area, -N_bilingual) 


```
#### convert to wide dataset
```{r}
table_city_wide = table_city_by_age_subgroup %>%
  # remove totals so they don't confuse pivot_wider
  select(-total) %>%
  pivot_wider(names_from = "age_collapsed", values_from = c("N_bilingual", "Percent"))
```

#### age 0-9
```{r}
table_city_0_9 = table_city_by_age_subgroup %>% 
  group_by(province, area) %>%
  summarize(Percent_age_0_to_9 = sum(N_bilingual)/sum(total))
```
#### combine all ages
```{r}
table_city_full= table_city_0_9 %>% full_join(table_city_wide) %>% select(-starts_with("N_bilin"))
```


### merge all tables for canada, provinces, areas

```{r}
combined_table = table_canada %>%
  full_join( table_city_full) %>%
  full_join(table_prov_combined) %>% 
  arrange(country,province, area) %>% 
  # convert all percent numbers to percent
  mutate_at(vars(starts_with("Percent")), ~round(.,3)*100) %>%
  relocate(c(province, area), .after = country)



```
## save data as csv for paper
```{r}
rio::export(combined_table, file = here("03_output/02_tables/Table1_bilingualism_rates.csv"))
```


## save geographical  data for maps
for the plot_map script
```{r}


## save percentages for maps (without the summary percentage for all of canada)

export(combined_table %>% filter(is.na(country)), here("03_output", "00_processed_data", "idata_geo_province.Rdata"))


```
# look at by-province rankings
for results section in paper
```{r}
table_prov_combined %>%  
  select(-area) %>%
  # round all numbers & convert to % for easier visibility
  mutate_at(vars(starts_with("Percent")), ~round(.,3)*100) %>% 
 # rank for 0-4 and 0-9
  mutate(rank_0_4 = rank(-Percent_age_0_to_4),
         rank_0_9 = rank(-Percent_age_0_to_9)) %>% 
  # arrange by 0-9 so discrepancies are more easily visible
  arrange(rank_0_9)
```






